---
title: "Constructing AML State-Space: mRNA (2018 Cohort)"
format: html
author: Danya Hassan
knitr:
  opts_chunk:
    echo: TRUE
    warning: FALSE
    error: TRUE
---

## Overview

This document constructs the AML transcriptomic state-space from the 2018 mRNA
cohort. The pipeline covers data loading and filtering, SVD-based dimensionality
reduction, rotation of the PC space so that control trajectories are flat, 
identification of critical points (c₁, c₂, c₃) via k-means clustering, and
construction of the quasi-potential double-well curve.

---

## 1. Setup

```{r}
#| label: libraries

library(ggplot2)
library(dplyr)
library(tidyr)
library(tidyverse)
library(gt)
library(SummarizedExperiment)
library(patchwork)
```

```{r}
#| label: load-data

mrna_2018 <- readRDS("AML_mRNA_2018_1_real.rds")
```

---

## 2. Data Exploration

```{r}
#| label: explore-se

head(mrna_2018)
assays(mrna_2018)
```

```{r}
#| label: sample-metadata

mrna18_info <- as.data.frame(colData(mrna_2018))
```

There are 14 mice in the 2018 AML experiment: 7 healthy (Ctrl) and 7 CM.

```{r}
#| label: mouse-counts

length(unique(mrna18_info$mouse_id))

mrna18_info %>%
  group_by(treatment) %>%
  summarise(n_mice = n_distinct(mouse_id))
```

Starting dimensions: 33,789 genes × 139 sample collections.

```{r}
#| label: data-dimensions

dim(assay(mrna_2018, "tpm"))
```

---

## 3. Data Filtering and Log-Scaling

Remove zero-expression rows, log₂-transform with a minimum-value offset, and
mean-center across Ctrl + CM samples to construct the state-space matrix.

```{r}
#| label: filtering-and-scaling

# Identify genes with any non-zero TPM expression
lowExpRM <- which(rowSums(data.matrix(assay(mrna_2018, "tpm"))) > 0)

# Minimum non-zero offsets (to avoid log(0))
minexp  <- min(data.matrix(assay(mrna_2018, "counts"))[
                 data.matrix(assay(mrna_2018, "counts")) > 0])
minabd  <- min(data.matrix(assay(mrna_2018, "tpm"))[
                 data.matrix(assay(mrna_2018, "tpm")) > 0])

# Log₂-scaled, mean-centered TPM matrix (samples × genes)
dat.mc <- scale(t(log2(data.matrix(assay(mrna_2018, "tpm")) + minabd)), scale = FALSE)
assays(mrna_2018)$almc <- t(dat.mc)   # abundance log mean-centered

# Log₂-scaled TPM (no centering)
dat.alog <- log2(data.matrix(assay(mrna_2018, "tpm")) + minabd)
assays(mrna_2018)$alog <- dat.alog    # abundance log

# Subset to Ctrl and CM samples for state-space construction
ldat  <- log2(data.matrix(assay(mrna_2018, "tpm")) + minabd)
cbsel <- which(colData(mrna_2018)$treatment == "Ctrl" |
               colData(mrna_2018)$treatment == "CM")
cb.mc <- rowMeans(ldat[, cbsel])

# Mean-center using Ctrl+CM means; drop low-expression genes
cb.almc <- sweep(ldat[lowExpRM, cbsel], 1, cb.mc[lowExpRM], FUN = "-")
```

---

## 4. SVD — State-Space Construction

Perform singular value decomposition on the centered expression matrix.
Rows of **U** are sample coordinates (PCs); columns of **V** are gene loadings
(eigengenes).

```{r}
#| label: svd

bsvd <- svd(t(cb.almc))
bV   <- bsvd$v
bU   <- bsvd$u
bD   <- bsvd$d
bDm  <- diag(bD)

rownames(bV) <- rowData(mrna_2018)$gene_name[lowExpRM]
rownames(bU) <- colnames(mrna_2018)[cbsel]
```

```{r}
#| label: svd-dimensions

print(paste("Dim of U:", paste(dim(bU), collapse = " x "),
            "| Dim of V:", paste(dim(bV), collapse = " x "),
            "| Dim of D:", length(bD)))
```

Variance explained by each PC:

```{r}
#| label: variance-explained

barplot(bD / sum(bD) * 100,
        col    = "black",
        border = NA,
        xlab   = "PC",
        ylab   = "% Variance Explained",
        main   = "Scree Plot — AML mRNA SVD")
```

---

## 5. Encode Timepoints as Numeric

Convert `T#` timepoint labels (e.g. `T2p5`) to numeric weeks.

```{r}
#| label: timepoint-encoding

mrna18_info <- mrna18_info %>%
  mutate(time_numeric = as.numeric(gsub("p", ".", gsub("T", "", timepoint))))

# Align back to the SummarizedExperiment by library_id
idx <- match(colData(mrna_2018)$library_id, mrna18_info$library_id)
colData(mrna_2018)$time_numeric <- mrna18_info$time_numeric[idx]
```

Build the initial sample dataframe with PC coordinates and metadata:

```{r}
#| label: build-sdf

sdf <- data.frame(
  PC1        = bU[, 1],
  PC2        = bU[, 2],
  PC3        = bU[, 3],
  PC4        = bU[, 4],
  PC5        = bU[, 5],
  PC6        = bU[, 6],
  PC7        = bU[, 7],
  Treatment  = colData(mrna_2018)$treatment[cbsel],
  Time       = colData(mrna_2018)$time_numeric[cbsel],
  mouse_id   = colData(mrna_2018)$mouse_id[cbsel],
  group      = colData(mrna_2018)$cohort[cbsel]
)

head(sdf)
```

---

## 6. Eigengenes for PC2

Eigengenes are the gene loadings (columns of **V**) for a given PC — the genes
that most strongly define the state-space axis.

```{r}
#| label: eigengenes-pc2

eigengene_PC2        <- bV[, 2]
names(eigengene_PC2) <- rownames(bV)

# Variance-weighted version (scales magnitude to reflect explained variance)
eigengene_pc2_scaled <- eigengene_PC2 * bD[2]
```

Sanity check — reconstructed PC2 scores should correlate perfectly with bU[,2]:

```{r}
#| label: eigengene-sanity-check

pc2_recon <- t(cb.almc) %*% eigengene_PC2
cor(pc2_recon, bU[, 2])   # expected: ±1
```

Top genes driving PC2 separation:

```{r}
#| label: top-pc2-genes

top_pc2_genes <- sort(abs(eigengene_PC2), decreasing = TRUE)
head(top_pc2_genes, 20)
```

Save eigengene table:

```{r}
#| label: save-eigengenes

eigengene_PC2_df <- data.frame(
  gene                 = names(eigengene_PC2),
  eigengene_PC2        = eigengene_PC2,
  eigengene_PC2_scaled = eigengene_pc2_scaled
)

write.csv(eigengene_PC2_df, "eigengenes_PC2_aml2018.csv", row.names = FALSE)
```

---

## 7. Rotating the State-Space

Rotate the PC1–PC2 plane so that the linear fit through **Ctrl** samples has
slope ≈ 0 (i.e., control trajectories are flat). PC2 is flipped if necessary
so that the leukemic state corresponds to negative PC2 values.

```{r}
#| label: flip-pc2

# Ensure CM samples are in the negative PC2 direction
if (mean(bU[, 2][which(sdf$Treatment == "CM")]) > 0) {
  bU[, 2] <- bU[, 2] * -1
}
```

Scan rotation angles to find the one that minimises the slope of the Ctrl fit:

```{r}
#| label: find-rotation-angle

test_deg <- seq(-3, 3, by = 0.001)
mlist    <- numeric(length(test_deg))

for (i in seq_along(test_deg)) {
  deg   <- test_deg[i]
  theta <- deg * pi / 180
  RotM  <- matrix(c(cos(theta), sin(theta), -sin(theta), cos(theta)),
                  nrow = 2, ncol = 2)
  rU_test <- bU[, c(1, 2)] %*% bDm[c(1, 2), c(1, 2)] %*% RotM

  r.exp.test <- data.frame(
    PC1       = rU_test[, 1],
    PC2       = rU_test[, 2],
    Treatment = colData(mrna_2018)$treatment[cbsel]
  )

  ctrl_fit <- lm(
    PC2 ~ PC1,
    data = r.exp.test[r.exp.test$Treatment == "Ctrl", ]
  )
  mlist[i] <- ctrl_fit$coefficients[[2]]
}

minM  <- which(abs(mlist) == min(abs(mlist)))
deg   <- test_deg[minM]
theta <- deg * pi / 180

cat("Optimal rotation angle:", deg, "degrees\n")
```

Apply the rotation:

```{r}
#| label: apply-rotation

RotM <- matrix(c(cos(theta), sin(theta), -sin(theta), cos(theta)),
               nrow = 2, ncol = 2)

rU <- bU[, c(1, 2)] %*% bDm[c(1, 2), c(1, 2)] %*% RotM   # scale by eigenvalues
rV <- bV[, c(1, 2)] %*% RotM
rownames(rV) <- rowData(mrna_2018[lowExpRM, ])$gene_name

r.exp.df <- data.frame(
  PC1        = rU[, 1],
  PC2        = rU[, 2],
  Treatment  = colData(mrna_2018)$treatment[cbsel],
  Time       = colData(mrna_2018)$time_numeric[cbsel],
  mouse_id   = colData(mrna_2018)$mouse_id[cbsel],
  Sex        = colData(mrna_2018)$sex[cbsel],
  group_name = colData(mrna_2018)$cohort[cbsel]
)

r.ctrl.fit <- lm(
  PC2 ~ PC1,
  data = r.exp.df[r.exp.df$Treatment == "Ctrl", ]
)

r.ctrl.fit
```

```{r}
#| label: plot-rotated-space
#| fig-width: 6
#| fig-height: 5

ggplot(r.exp.df, aes(x = PC1, y = PC2, color = Treatment)) +
  geom_point(size = 2) +
  geom_abline(
    intercept = r.ctrl.fit$coefficients[[1]],
    slope     = r.ctrl.fit$coefficients[[2]],
    linetype  = "dashed",
    color     = "black"
  ) +
  theme_bw(base_size = 16) +
  ggtitle("AML mRNA — Rotated State-Space (PC1 vs PC2)")
```

```{r}
#| label: plot-pc2-vs-time
#| fig-width: 7
#| fig-height: 5

ggplot(r.exp.df, aes(x = Time, y = PC2, color = Treatment, group = mouse_id)) +
  geom_point() +
  geom_line() +
  theme_bw(base_size = 16) +
  xlab("Time (weeks)") +
  ylab("PC2") +
  ggtitle("AML mRNA — PC2 vs Time")
```

---

## 8. Project All Samples into State-Space

Map every sample (Ctrl and CM) to its rotated PC2 coordinate. Samples from
treatment groups not used in SVD construction would be projected here via `rV`.

```{r}
#| label: project-all-samples

cbcol     <- colData(mrna_2018)[cbsel, ]
aml_space <- c()
samp_test <- c()

for (i in seq_len(dim(colData(mrna_2018))[1])) {
  cursamp   <- colData(mrna_2018)$sample_id[i]
  samp_test <- c(samp_test, cursamp)

  if (colData(mrna_2018)$treatment[i] %in% c("Ctrl", "CM")) {
    m <- which(cbcol$sample_id == cursamp)
    if (length(m) != 1) {
      message("ERROR: matching sample ", cursamp)
    } else {
      aml_space <- c(aml_space, rU[m, 2])
    }
  } else {
    message("ERROR: unrecognized treatment for sample ", cursamp)
  }
}

tdf <- data.frame(
  AML_space.calc = aml_space,
  mouse_id       = colData(mrna_2018)$mouse_id,
  sample_id      = samp_test,
  Time           = colData(mrna_2018)$time_numeric,
  group_name     = colData(mrna_2018)$treatment
)

head(tdf)
```

```{r}
#| label: plot-all-trajectories
#| fig-width: 8
#| fig-height: 5

ggplot(tdf, aes(x = Time, y = AML_space.calc,
                color = factor(mouse_id), group = mouse_id)) +
  geom_point() +
  geom_line() +
  theme_bw(base_size = 16) +
  ggtitle("All Mice — AML State-Space Trajectories") +
  theme(plot.title = element_text(size = 13))
```

---

## 9. Outlier Sample Filtering

Based on visual inspection of trajectories, the following samples are excluded
before downstream analysis:

- Mouse **3368**: hovers near the healthy state throughout; not a leukemic trajectory.
- Mouse **3336**, week 8.0: anomalous late recovery point.
- Mouse **3357**, all timepoints after week 3.0: post-week-3 behavior is atypical.

```{r}
#| label: filter-outliers

final_tdf <- tdf %>%
  filter(
    mouse_id != 3368,
    !(mouse_id == 3336 & Time == 8.0),
    !(mouse_id == 3357 & Time > 3.0)
  )

# Verify no excluded samples remain
violations <- final_tdf %>%
  filter(
    mouse_id == 3368 |
    (mouse_id == 3336 & Time == 8.0) |
    (mouse_id == 3357 & Time > 3.0)
  )
stopifnot(nrow(violations) == 0)

cat("Samples remaining after filtering:", nrow(final_tdf), "\n")
```

```{r}
#| label: plot-filtered-trajectories
#| fig-width: 8
#| fig-height: 5

ggplot(final_tdf, aes(x = Time, y = AML_space.calc,
                      color = factor(group_name), group = mouse_id)) +
  geom_point() +
  geom_line() +
  theme_bw(base_size = 16) +
  ggtitle("Filtered Mice — AML State-Space Trajectories") +
  theme(plot.title = element_text(size = 13))
```

---

## 10. Merge with Sample Metadata and Save

```{r}
#| label: merge-metadata

mrna_2018_ss <- final_tdf %>%
  left_join(mrna18_info, by = "sample_id")

write.csv(mrna_2018_ss, "AML_2018_StateSpace_Mice.csv", row.names = FALSE)
```

---

## 11. K-Means Clustering Along the AML Axis

Cluster CM samples into three states along PC2 to identify empirical critical
points of the quasi-potential.

- **K1** — perturbed hematopoiesis (near-healthy)
- **K2** — transition region
- **K3** — deep leukemia

```{r}
#| label: load-filtered-data

# Re-load if continuing from a saved file
final_tdf <- read.csv("final_aml_mrna_state_space.csv")
```

```{r}
#| label: kmeans-clustering

tdf_cm <- final_tdf %>%
  filter(group_name == "CM")

set.seed(56)
cm_kmeans     <- kmeans(tdf_cm$AML_space.calc, centers = 3)
tdf_cm$cluster <- factor(cm_kmeans$cluster)

cat("K-means cluster centers:\n")
print(cm_kmeans$centers)
```

Order and label clusters along the AML progression axis:

```{r}
#| label: label-clusters

cluster_centers <- tdf_cm %>%
  group_by(cluster) %>%
  summarise(center = mean(AML_space.calc)) %>%
  arrange(center)

cluster_centers

cluster_map <- tibble(
  cluster = factor(c(1, 2, 3)),  #Adjust so order goes from most positive centers (near health) --> negative (disease)
  state   = c("K1_perturbed_hematopoiesis",
              "K2_transition_region",
              "K3_deep_leukemia")
)

tdf_cm <- tdf_cm %>%
  left_join(cluster_map, by = "cluster")
```

```{r}
#| label: plot-clusters
#| fig-width: 5
#| fig-height: 4

ggplot(tdf_cm, aes(x = cluster, y = AML_space.calc)) +
  geom_boxplot() +
  theme_bw(base_size = 14) +
  ggtitle("AML State Values by K-Means Cluster")
```

```{r}
#| label: save-clustered-cm

write.csv(tdf_cm, "cm2018_sp_kmeansclusters.csv", row.names = FALSE)
```

---

## 12. Estimate Critical Points c₁, c₂, c₃

```{r}
#| label: critical-points

# c1* : grand mean of all Ctrl samples (healthy attractor)
c1_star <- final_tdf %>%
  filter(group_name == "Ctrl") %>%
  summarise(val = mean(AML_space.calc)) %>%
  pull(val)

# c1 : centroid of K1 (CM near-health cluster)
c1 <- tdf_cm %>%
  filter(state == "K1_perturbed_hematopoiesis") %>%
  summarise(val = mean(AML_space.calc)) %>%
  pull(val)

# c3 : centroid of K3 (deep leukemia cluster)
c3 <- tdf_cm %>%
  filter(state == "K3_deep_leukemia") %>%
  summarise(val = mean(AML_space.calc)) %>%
  pull(val)

# c2 : upper bound of K2 (transition region)
# The unstable maximum is least-sampled; its upper boundary is the best
# empirical estimate because a higher Boltzmann ratio is expected here.
c2_range <- tdf_cm %>%
  filter(state == "K2_transition_region") %>%
  summarise(c2_lower = min(AML_space.calc),
            c2_upper = max(AML_space.calc))

c2_upper_hat <- c2_range$c2_upper

cat("Critical points:\n")
cat(sprintf("  c1*  (healthy, Ctrl centroid) : %.4f\n", c1_star))
cat(sprintf("  c1   (K1 CM centroid)         : %.4f\n", c1))
cat(sprintf("  c2   (K2 upper bound)         : %.4f\n", c2_upper_hat))
cat(sprintf("  c3   (K3 CM centroid)         : %.4f\n", c3))

# Ordering sanity checks
stopifnot(c1 > c2_upper_hat, c2_upper_hat > c3)
```

```{r}
#| label: critical-points-table

critical_points <- data.frame(
  point     = c("c1", "c2", "c3"),
  AML_space = c(c1, c2_upper_hat, c3),
  stability = c(
    "Stable minimum (perturbed hematopoiesis)",
    "Unstable maximum (transition state)",
    "Stable minimum (leukemia)"
  )
)

critical_points
```

---

## 13. Quasi-Potential Construction

The quasi-potential is a quartic polynomial whose derivative factors as:

$$\frac{dU_p}{dx} = \alpha (x - c_1)(x - c_2)(x - c_3)$$

Integrating gives:

$$U_p(x) = \alpha \left[ \frac{x^4}{4} - \frac{(c_1+c_2+c_3)x^3}{3} + \frac{(c_1 c_2 + c_1 c_3 + c_2 c_3)x^2}{2} - c_1 c_2 c_3 \cdot x \right]$$

```{r}
#| label: define-potential

alpha <- 4.85e-8   # scaling parameter (chosen to match diffusion fit)

make_potential <- function(x, c1, c2, c3, alpha) {
  alpha * (
    x^4 / 4 -
    (c1 + c2 + c3) * x^3 / 3 +
    (c1*c2 + c1*c3 + c2*c3) * x^2 / 2 -
    (c1*c2*c3) * x
  )
}
```

### 13a. Family of quasi-potentials across candidate c₂ values

```{r}
#| label: quasi-potential-family
#| fig-width: 8
#| fig-height: 5

x_grid           <- seq(c3 - 50, c1 + 50, length.out = 500)
c2_range_c1_c3   <- seq(c1, c3, length.out = 200)

U_family <- lapply(c2_range_c1_c3, function(c2) {
  make_potential(x_grid, c1, c2, c3, alpha)
})

U_family_idx    <- round(seq(1, length(c2_range_c1_c3), length.out = 10))
U_c2_values_df  <- map_dfr(U_family_idx, function(i) {
  tibble(x  = x_grid,
         U  = U_family[[i]],
         c2 = c2_range_c1_c3[i])
})

ggplot(U_c2_values_df, aes(x = x, y = U, group = c2)) +
  geom_line(color = "gray50", linewidth = 0.7, alpha = 0.8) +
  scale_x_reverse() +
  geom_vline(xintercept = c(c1, c3),
             linetype = "dashed", linewidth = 0.6) +
  geom_vline(xintercept = c2_upper_hat,
             linetype = "dashed", color = "red") +
  annotate("text", x = c1, y = min(U_c2_values_df$U),
           label = "K1", vjust = 1.0, size = 4.5) +
  annotate("text", x = c3, y = min(U_c2_values_df$U),
           label = "K3", vjust = 1.0, size = 4.5) +
  annotate("text", x = c2_upper_hat, y = min(U_c2_values_df$U),
           label = "K2 upper", vjust = 1.0, size = 4.5) +
  labs(x  = "AML state-space (PC2)",
       y  = expression(U[p](x)),
       title = expression("Quasi-potentials for varying " * c[2])) +
  theme_classic(base_size = 14)

write.csv(U_c2_values_df, "quasipotential_varying_c2.csv", row.names = FALSE)
```

### 13b. Final quasi-potential — c₂ = K2 upper bound

```{r}
#| label: final-potential-k2max
#| fig-width: 7
#| fig-height: 5

c2 <- c2_upper_hat

a4 <- alpha / 4
a3 <- -alpha * (c1 + c2 + c3) / 3
a2 <-  alpha * (c1*c2 + c1*c3 + c2*c3) / 2
a1 <- -alpha * (c1*c2*c3)
a0 <- 0

U_poly <- function(x) a4*x^4 + a3*x^3 + a2*x^2 + a1*x + a0

U_vals                     <- U_poly(x_grid)
U_vals_final               <- U_vals - min(U_vals)
final_aml_potential_k2max  <- data.frame(AML_PC2_Location = x_grid,
                                         U_vals           = U_vals_final)

ggplot(final_aml_potential_k2max, aes(x = AML_PC2_Location, y = U_vals)) +
  geom_line(color = "darkblue", linewidth = 0.7) +
  scale_x_reverse() +
  geom_vline(xintercept = c(c1, c2, c3),
             linetype = "dashed", linewidth = 0.6,
             color = c("black", "red", "darkred")) +
  labs(x        = "AML state-space (PC2)",
       y        = expression(U[p](x)),
       title    = "Double-Well AML mRNA Quasi-Potential",
       subtitle = expression(c[2] * " = K"[2] * " upper bound")) +
  theme_classic(base_size = 14)

write.csv(final_aml_potential_k2max, "aml_qp_dw_k2max.csv", row.names = FALSE)
```

### 13c. Final quasi-potential — c₂ = K2 centroid

```{r}
#| label: final-potential-k2centroid
#| fig-width: 7
#| fig-height: 5

k2_centroid <- cm_kmeans$centers[2]

a4 <- alpha / 4
a3 <- -alpha * (c1 + k2_centroid + c3) / 3
a2 <-  alpha * (c1*k2_centroid + c1*c3 + k2_centroid*c3) / 2
a1 <- -alpha * (c1*k2_centroid*c3)
a0 <- 0

U_poly_centroid <- function(x) a4*x^4 + a3*x^3 + a2*x^2 + a1*x + a0

U_vals_centroid              <- U_poly_centroid(x_grid)
U_vals_centroid_final        <- U_vals_centroid - min(U_vals_centroid)
final_aml_potential_k2cent   <- data.frame(AML_PC2_Location = x_grid,
                                           U_vals           = U_vals_centroid_final)

ggplot(final_aml_potential_k2cent, aes(x = AML_PC2_Location, y = U_vals)) +
  geom_line(color = "darkblue", linewidth = 0.7) +
  scale_x_reverse() +
  geom_vline(xintercept = c(c1, k2_centroid, c3),
             linetype = "dashed", linewidth = 0.6,
             color = c("black", "red", "darkred")) +
  labs(x        = "AML state-space (PC2)",
       y        = expression(U[p](x)),
       title    = "Double-Well AML mRNA Quasi-Potential",
       subtitle = expression(c[2] * " = K"[2] * " centroid")) +
  theme_classic(base_size = 14)

write.csv(final_aml_potential_k2cent, "amlss_quasipotential_dwcurve.csv",
          row.names = FALSE)
```

---

## 14. Curvature Check

Verify that c₁ and c₃ are local minima (U'' > 0) and c₂ is a local maximum
(U'' < 0):

```{r}
#| label: curvature-check

U_dd <- function(x, c1, c2, c3, alpha) {
  alpha * (3*x^2 - 2*(c1 + c2 + c3)*x + (c1*c2 + c1*c3 + c2*c3))
}

cat("U''(c1) =", U_dd(c1, c1, c2_upper_hat, c3, alpha),
    "  [should be > 0]\n")
cat("U''(c2) =", U_dd(c2_upper_hat, c1, c2_upper_hat, c3, alpha),
    "  [should be < 0]\n")
cat("U''(c3) =", U_dd(c3, c1, c2_upper_hat, c3, alpha),
    "  [should be > 0]\n")
```

---

## 15. Boltzmann Ratio Across Candidate c₂ Values

The Boltzmann ratio BR = Pr(c₃) / Pr(c₁) should be > 1 (samples are more
likely to be in the leukemic basin) and increases toward the leftmost (c₁-side)
boundary of K2. This is used to confirm that the upper bound of K2 is a
principled choice for c₂.

```{r}
#| label: boltzmann-ratio
#| fig-width: 7
#| fig-height: 5

kT <- 1   # Boltzmann constant × temperature; set to 1 for shape analysis

BR_vals <- sapply(c2_range_c1_c3, function(c2_cand) {
  U_c1 <- make_potential(c1, c1, c2_cand, c3, alpha)
  U_c3 <- make_potential(c3, c1, c2_cand, c3, alpha)
  exp(-(U_c3 - U_c1) / kT)
})

BR_df <- data.frame(c2 = c2_range_c1_c3, BR = BR_vals)

ggplot(BR_df, aes(x = c2, y = BR)) +
  geom_line(linewidth = 1.1, color = "black") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  scale_x_reverse() +
  xlim(c1, c3) +
  labs(x     = expression(c[2] ~ "(candidate unstable state)"),
       y     = "Boltzmann Ratio  Pr(c₃) / Pr(c₁)",
       title = "Boltzmann Ratio Across Candidate c₂ Values") +
  theme_classic(base_size = 14)
```

---

## 16. Monte Carlo Simulation Setup

> **Next step:** Use the finalized quasi-potential to run Metropolis–Hastings
> Monte Carlo simulations, discretize trajectories into the three macro-states
> (K1, K2, K3), and estimate the 3×3 Markov transition matrix.

*Monte Carlo simulation and Markov matrix construction are implemented in a
separate document.*

---

## Session Info

```{r}
#| label: session-info

sessionInfo()
```
