---
title: "AML Monte Carlo Simulations — c₃ Predictions and Markov Analysis"
format: html
author: Danya Hassan
knitr:
  opts_chunk:
    echo: TRUE
    warning: FALSE
    error: TRUE
---

## Overview

This document runs the calibrated Monte Carlo (MC) simulation on the rescaled
AML quasi-potential landscape to produce time-to-leukemia (c₃) predictions at
both the **per-mouse** and **population** levels. It also constructs τ-lagged
Markov transition matrices, computes mean first-passage times (MFPT) analytically,
and validates population-level predictions via bootstrap test.

**Dependencies:** Objects produced in `aml2018_mc_parameterization.qmd` are
assumed to be available — specifically: `x_seq`, `U_seq`, `states`, `n_pts`,
`propose_index()`, `run_fpt()`, `c1_idx`, `step_to_month`, `D`,
`step_sd_coord`, `max_step_coord`, `n_steps`, `boundary_1`, `boundary_2`,
`obs_all`, `calibration_mice`, `validation_mice`, and `aml_ss`.

---

## 1. Setup

```{r}
#| label: libraries

library(ggplot2)
library(dplyr)
library(tidyr)
library(tidyverse)
library(patchwork)
library(readr)
library(ggrepel)
library(gganimate)
library(gifski)
```

```{r}
#| label: load-data

aml_ss    <- read.csv("final_aml_mrna_state_space.csv")
aml_qp_dw <- read.csv("aml_qp_dw_k2max.csv")

# Critical points and simulation parameters (reproduced from parameterization doc)
c1 <- 14.89216
c2 <- -40.5698567718201
c3 <- -299.0184

boundary_1 <- (c1 + c2) / 2
boundary_2 <- -250

D              <- 0.45
n_steps        <- 50000
step_sd_coord  <- 15
max_step_coord <- 22.5

calibration_mice <- c(3334, 3357, 3336, 3341)
validation_mice  <- c(3370, 3338)

obs_all <- data.frame(
  mouse_id      = c(3334, 3336, 3338, 3341, 3357, 3370),
  observed_time = c(6.5,  7.0,  3.0,  6.0,  3.0,  9.0)
)

mean_obs_months_calib <- mean(obs_all$observed_time[
  obs_all$mouse_id %in% calibration_mice])
```

```{r}
#| label: build-landscape

# Sort landscape and assign discrete states
landscape <- aml_qp_dw[order(aml_qp_dw$AML_PC2_Location), ]
x_seq     <- landscape$AML_PC2_Location
U_seq     <- landscape$U_vals
states    <- landscape$state
n_pts     <- length(x_seq)

# Rescale U to 0.99 A.U. barrier (matching reference paper)
target_barrier   <- 0.99
current_barrier  <- U_seq[which.min(abs(x_seq - boundary_1))] -
                    min(U_seq[states == 1L])
u_rescale_factor <- target_barrier / current_barrier
U_seq            <- U_seq * u_rescale_factor

# State assignments
aml_qp_dw$state <- cut(
  aml_qp_dw$AML_PC2_Location,
  breaks = c(-Inf, boundary_2, boundary_1, Inf),
  labels = c("3", "2", "1")
) |> as.character() |> as.integer()

aml_ss$state <- cut(
  aml_ss$AML_space.calc,
  breaks = c(-Inf, boundary_2, boundary_1, Inf),
  labels = c("3", "2", "1")
) |> as.character() |> as.integer()

c1_idx <- which(states == 1L)
```

```{r}
#| label: simulation-functions

# Metropolis-Hastings proposal with reflecting boundaries
propose_index <- function(cur_idx) {
  cur_coord  <- x_seq[cur_idx]
  delta      <- rnorm(1, 0, step_sd_coord)
  delta      <- max(min(delta, max_step_coord), -max_step_coord)
  if (abs(delta) < 1e-6)
    delta <- sample(c(-1, 1), 1) * step_sd_coord * 0.1
  prop_coord <- cur_coord + delta
  x_min <- x_seq[1]; x_max <- x_seq[n_pts]
  if (prop_coord < x_min) prop_coord <- 2 * x_min - prop_coord
  if (prop_coord > x_max) prop_coord <- 2 * x_max - prop_coord
  which.min(abs(x_seq - prop_coord))
}

# First-passage time: returns step count when trajectory first enters State 3
run_fpt <- function(start_idx, n_steps, D) {
  cur <- start_idx
  for (t in seq_len(n_steps)) {
    prop   <- propose_index(cur)
    deltaU <- U_seq[prop] - U_seq[cur]
    if (runif(1) < min(1, exp(-deltaU / D))) cur <- prop
    if (states[cur] == 3L) return(t)
  }
  return(NA_integer_)
}

# Binned mode helper for right-skewed FPT distributions
binned_mode <- function(x, n_bins = 40) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  h <- hist(x, breaks = n_bins, plot = FALSE)
  h$mids[which.max(h$counts)]
}
```

```{r}
#| label: calibrate-step-to-month

set.seed(42)
calib_steps <- vapply(seq_len(200), function(i)
  run_fpt(sample(c1_idx, 1), n_steps, D), integer(1))

mean_calib_steps <- mean(calib_steps, na.rm = TRUE)
step_to_month    <- mean_obs_months_calib / mean_calib_steps

cat(sprintf("Mean calibration steps: %.1f\nstep_to_month:          %.6f\n",
            mean_calib_steps, step_to_month))
```

---

## 2. Per-Mouse FPT Predictions

Each mouse starts at its first observed transcriptomic coordinate. 700
independent trajectories are simulated per mouse.

```{r}
#| label: per-mouse-starting-coords

start_coords <- aml_ss %>%
  filter(group_name == "CM") %>%
  arrange(mouse_id, Time) %>%
  group_by(mouse_id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  select(mouse_id, AML_space.calc)

cor_val <- cor(
  start_coords$AML_space.calc,
  obs_all$observed_time[match(start_coords$mouse_id, obs_all$mouse_id)],
  use = "complete.obs"
)

start_coords %>%
  left_join(obs_all, by = "mouse_id") %>%
  mutate(role = if_else(mouse_id %in% calibration_mice,
                        "calibration", "validation")) %>%
  arrange(observed_time)

cat(sprintf("\nCorrelation (start coord vs observed time): %.3f\n", cor_val))
```

```{r}
#| label: run-per-mouse-fpt

n_pred_reps <- 700
cm_mice     <- sort(unique(start_coords$mouse_id))

predict_mouse <- function(mouse_id_val, n_reps = n_pred_reps, D_val = D) {
  start_coord <- aml_ss %>%
    filter(mouse_id == mouse_id_val) %>%
    arrange(Time) %>%
    dplyr::slice(1) %>%
    pull(AML_space.calc)

  start_idx <- which.min(abs(x_seq - start_coord))

  cat(sprintf("  Mouse %s | coord = %.3f | state = %d\n",
              mouse_id_val, start_coord, states[start_idx]))

  steps_vec  <- vapply(seq_len(n_reps),
                       function(i) run_fpt(start_idx, n_steps, D_val),
                       integer(1))
  months_vec <- steps_vec * step_to_month

  list(
    mouse_id         = mouse_id_val,
    predicted_mean   = mean(months_vec,   na.rm = TRUE),
    predicted_median = median(months_vec, na.rm = TRUE),
    predicted_mode   = binned_mode(months_vec),
    predicted_sd     = sd(months_vec,     na.rm = TRUE),
    pct_reached      = mean(!is.na(steps_vec)),
    months_vec       = months_vec,
    steps_vec        = steps_vec
  )
}

cat("Predicting all CM mice...\n")
results_list <- lapply(cm_mice, function(m) {
  cat(sprintf("Mouse %s:\n", m)); predict_mouse(m)
})
```

```{r}
#| label: per-mouse-summary

summary_df <- bind_rows(lapply(results_list, function(r) {
  data.frame(
    mouse_id         = r$mouse_id,
    predicted_mean   = r$predicted_mean,
    predicted_median = r$predicted_median,
    predicted_mode   = r$predicted_mode,
    predicted_sd     = r$predicted_sd,
    pct_reached      = r$pct_reached,
    stringsAsFactors = FALSE
  )
})) %>%
  left_join(obs_all, by = "mouse_id") %>%
  mutate(
    error_mean   = predicted_mean   - observed_time,
    error_median = predicted_median - observed_time,
    error_mode   = predicted_mode   - observed_time,
    calib_flag   = if_else(mouse_id %in% calibration_mice,
                           "calibration", "validation")
  )

cat("\n=== Per-Mouse Prediction Summary ===\n")
print(summary_df[, c("mouse_id", "calib_flag", "observed_time",
                     "predicted_median", "predicted_mean",
                     "error_mean", "error_median")])
```

```{r}
#| label: build-sim-long

# Long-format simulation table: one row per trajectory per mouse
sim_long <- bind_rows(lapply(results_list, function(r) {
  data.frame(mouse_id   = r$mouse_id,
             sim_months = r$months_vec,
             sim_steps  = r$steps_vec,
             stringsAsFactors = FALSE)
}))

cat(sprintf("sim_long: %d rows × %d cols  (%d mice × %d reps)\n",
            nrow(sim_long), ncol(sim_long),
            length(cm_mice), n_pred_reps))
```

---

## 3. Per-Mouse Prediction Plots

```{r}
#| label: plot-fpt-histograms
#| fig-width: 9
#| fig-height: 7

months_long <- bind_rows(lapply(results_list, function(r) {
  data.frame(mouse_id = r$mouse_id,
             months   = r$months_vec[!is.na(r$months_vec)])
})) %>% left_join(obs_all, by = "mouse_id")

p1 <- ggplot(months_long, aes(x = months)) +
  geom_histogram(bins = 40, fill = "#7F77DD",
                 color = "white", linewidth = 0.2) +
  geom_vline(aes(xintercept = observed_time),
             color = "#D85A30", linewidth = 0.9, linetype = "dashed") +
  geom_vline(data = summary_df, aes(xintercept = predicted_mean),
             color = "#1D9E75", linewidth = 0.9) +
  facet_wrap(~ mouse_id, scales = "free_x") +
  labs(title    = "Predicted time-to-leukemia distribution per mouse",
       subtitle = "Green solid = predicted mean | Red dashed = observed",
       x = "Predicted time (months)", y = "Count") +
  theme_bw(base_size = 13)

print(p1)
```

```{r}
#| label: plot-predicted-vs-observed
#| fig-width: 6
#| fig-height: 5

summary_long <- summary_df %>%
  select(mouse_id, calib_flag, observed_time,
         predicted_median, predicted_mode) %>%
  pivot_longer(cols = c(predicted_median, predicted_mode),
               names_to = "statistic", values_to = "predicted")

p2 <- ggplot(summary_long,
    aes(x = observed_time, y = predicted,
        color = calib_flag, shape = statistic)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey50") +
  geom_point(size = 4, alpha = 0.85) +
  scale_color_manual(values = c("calibration" = "#3B8BD4",
                                "validation"  = "#D85A30")) +
  scale_shape_manual(values = c("predicted_median" = 16,
                                "predicted_mode"   = 17)) +
  labs(title    = "Predicted vs observed time to leukemia (MC simulation)",
       subtitle = "Dashed = perfect prediction",
       x = "Observed time (months)", y = "Predicted time (months)",
       color = "Mouse role", shape = "Summary stat") +
  theme_bw(base_size = 13)

print(p2)
```

---

## 4. Distribution Diagnostics

Cross-check that `summary_df` means and medians match the raw simulation vectors,
and inspect the right-skew structure of FPT distributions.

```{r}
#| label: simulation-verification

sim_check <- sim_long %>%
  group_by(mouse_id) %>%
  summarise(
    mean_from_sims   = mean(sim_months,   na.rm = TRUE),
    median_from_sims = median(sim_months, na.rm = TRUE),
    n_reps           = n(),
    pct_na           = mean(is.na(sim_months)),
    .groups = "drop"
  ) %>%
  left_join(summary_df %>%
              select(mouse_id, predicted_mean, predicted_median,
                     observed_time),
            by = "mouse_id") %>%
  mutate(mean_diff   = mean_from_sims - predicted_mean,
         median_diff = median_from_sims - predicted_median)

cat("=== Simulation check vs summary_df (diffs should be ~0) ===\n")
print(sim_check)
```

```{r}
#| label: distribution-diagnostics

cat("=== Detailed FPT distribution diagnostics per mouse ===\n\n")

for (r in results_list) {
  x <- r$months_vec[!is.na(r$months_vec)]
  n <- length(x)

  mean_base    <- mean(x)
  h            <- hist(x, breaks = 40, plot = FALSE)
  mean_from_hist <- sum(h$mids * h$counts) / sum(h$counts)
  skew           <- mean((x - mean(x))^3) / sd(x)^3
  top5_contrib   <- sum(x[x > quantile(x, 0.95)]) / sum(x)

  cat(sprintf("Mouse %s (obs = %.1f months):\n", r$mouse_id,
      obs_all$observed_time[obs_all$mouse_id == r$mouse_id]))
  cat(sprintf("  n sims:                  %d\n", n))
  cat(sprintf("  mean (base R):           %.4f\n", mean_base))
  cat(sprintf("  mean (from histogram):   %.4f  [diff: %.6f]\n",
              mean_from_hist, abs(mean_base - mean_from_hist)))
  cat(sprintf("  Q25 / median / Q75:      %.2f / %.2f / %.2f\n",
              quantile(x, 0.25), median(x), quantile(x, 0.75)))
  cat(sprintf("  Q90 / Q95 / max:         %.2f / %.2f / %.2f\n",
              quantile(x, 0.90), quantile(x, 0.95), max(x)))
  cat(sprintf("  Skewness:                %.3f\n", skew))
  cat(sprintf("  Top 5%% contributes       %.1f%% of total sum\n\n",
              top5_contrib * 100))
}
```

---

## 5. Quasi-Potential Landscape Plot

```{r}
#| label: plot-landscape
#| fig-width: 7
#| fig-height: 4

ggplot(aml_qp_dw, aes(x = AML_PC2_Location, y = U_vals,
                       color = factor(state))) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = c(boundary_1, boundary_2),
             linetype = "dashed", color = "grey40") +
  scale_color_manual(values = c("1" = "#3B8BD4",
                                "2" = "#EF9F27",
                                "3" = "#D85A30"),
                     labels = c("State 1 (healthy)",
                                "State 2 (transition)",
                                "State 3 (leukemia)")) +
  scale_x_reverse() +
  labs(title    = "Quasi-potential landscape with state boundaries",
       subtitle = sprintf("Original U scale; simulation used U × %.2f (0.99 A.U. barrier)",
                          u_rescale_factor),
       x = "AML PC2 (state-space location)",
       y = "Potential energy U (original scale)",
       color = "State") +
  theme_bw(base_size = 13)
```

---

## 6. Eigengene Loadings — Drivers of State Transitions

Variance-scaled PC2 eigengene loadings reveal which genes drive transcriptomic
movement along the AML reaction coordinate. Confounding gene classes are removed
before visualization.

```{r}
#| label: load-eigengenes

eigengenes_df <- read.csv("eigengenes_PC2_aml2018.csv")
```

```{r}
#| label: filter-confounders

confound_patterns <- c(
  "^Gm",                            # predicted/uncharacterized genes
  "^A[0-9]", "^[0-9]",             # number-starting uncharacterized loci
  "Eif2s3y", "Ddx3y", "Uty",       # Y-chromosome (sex confound)
  "Kdm5d", "Ssty1", "Ssty2", "Sry", "Rbmy",
  "^Igh", "^Igk", "^Igl", "^Jchain",  # immunoglobulin (B-cell confound)
  "^Traj", "^Trbj", "^Trdj", "^Trgj", # T-cell receptor
  "^Rps", "^Rpl", "^Mrps", "^Mrpl",   # ribosomal (technical)
  "^mt-",                              # mitochondrial (technical/stress)
  "^Mcpt", "^Cma",                     # mast cell (cell type confound)
  "^Hist", "^H2a", "^H2b", "^H3",      # histones (cell cycle)
  "^H4", "^H1",
  "^Snor", "^Snhg",                    # small nucleolar/nuclear RNAs
  "Rik$",                              # uncharacterized Rik loci
  "^ENSMUSG",                          # unannotated Ensembl IDs
  "^Sox.*os", "^Platr", "Tas2r"        # antisense RNAs, placenta lncRNAs, taste receptors
)

pattern_combined <- paste(confound_patterns, collapse = "|")

eigengenes_clean <- eigengenes_df %>%
  filter(!grepl(pattern_combined, gene, perl = TRUE))

cat(sprintf("Before filtering: %d genes\n", nrow(eigengenes_df)))
cat(sprintf("After filtering:  %d genes\n", nrow(eigengenes_clean)))
cat(sprintf("Removed:          %d genes\n",
            nrow(eigengenes_df) - nrow(eigengenes_clean)))
```

```{r}
#| label: plot-eigengenes
#| fig-width: 8
#| fig-height: 6

# Highlighted genes of biological interest
highlight_genes <- c("Wnt2", "Bex4", "Cited1", "Rprm",
                     "Dbx2", "Prkd1", "Calr4", "Pcp4", "Plg")

top_pos <- eigengenes_clean %>%
  filter(eigengene_PC2_scaled > 0) %>%
  arrange(desc(eigengene_PC2_scaled)) %>%
  slice_head(n = 15)

top_neg <- eigengenes_clean %>%
  filter(eigengene_PC2_scaled < 0) %>%
  arrange(eigengene_PC2_scaled) %>%
  slice_head(n = 15)

plot_df <- bind_rows(top_pos, top_neg) %>%
  mutate(
    direction = if_else(eigengene_PC2 > 0,
                        "Toward healthy (State 1)",
                        "Toward leukemia (State 3)"),
    gene      = as.character(gene),
    label     = if_else(gene %in% highlight_genes, gene, ""),
    alpha_val = if_else(gene %in% highlight_genes, 1.0, 0.65)
  )

ggplot(plot_df,
    aes(x = eigengene_PC2_scaled,
        y = reorder(gene, eigengene_PC2_scaled),
        fill = direction, alpha = alpha_val)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = label,
                hjust = if_else(eigengene_PC2_scaled > 0, -0.1, 1.1)),
            size = 2.4, fontface = "bold", color = "seagreen") +
  geom_vline(xintercept = 0, linewidth = 0.5, color = "grey40") +
  scale_fill_manual(values = c("Toward healthy (State 1)" = "#185FA5",
                               "Toward leukemia (State 3)"  = "#D85A30")) +
  scale_alpha_identity() +
  scale_x_continuous(limits = c(-70, 35), expand = c(0.01, 0.01)) +
  labs(title = "PC2 eigengene loadings driving AML state transitions",
       x     = "PC2 loading (variance-scaled)",
       y     = NULL,
       fill  = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position    = "bottom",
        panel.grid.minor   = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.text.y        = element_blank(),
        axis.ticks.y       = element_blank())
```

---

## 7. Markov Transition Matrices

τ-lagged transition matrices are estimated by running long trajectories and
counting state-to-state transitions separated by lag τ. Three lag values are
tested (τ = 40, 50, 60) to confirm stability.

```{r}
#| label: markov-functions

n_markov_trajs <- 100
n_markov_steps <- 50000
n_macro        <- 3

# General builder — starts from a uniform sample of State 1
build_transition_matrix <- function(tau, D_val = D, seed_val = 67) {
  set.seed(seed_val)
  counts <- matrix(0L, n_macro, n_macro)

  for (tr in seq_len(n_markov_trajs)) {
    cur        <- sample(which(states == 1L), 1)
    traj_macro <- integer(n_markov_steps)
    for (t in seq_len(n_markov_steps)) {
      prop   <- propose_index(cur)
      deltaU <- U_seq[prop] - U_seq[cur]
      if (runif(1) < min(1, exp(-deltaU / D_val))) cur <- prop
      traj_macro[t] <- states[cur]
    }
    for (t in seq_len(n_markov_steps - tau)) {
      i <- traj_macro[t]; j <- traj_macro[t + tau]
      counts[i, j] <- counts[i, j] + 1L
    }
  }

  P <- counts / rowSums(counts)
  rownames(P) <- colnames(P) <- paste0("State ", 1:n_macro)
  P
}

# Per-mouse builder — every trajectory resets to the mouse's starting coordinate
build_transition_matrix_startcoord <- function(tau, D_val = D,
                                               seed_val = 67,
                                               start_coord) {
  set.seed(seed_val)
  start_idx <- which.min(abs(x_seq - start_coord))
  counts    <- matrix(0L, n_macro, n_macro)

  for (tr in seq_len(n_markov_trajs)) {
    cur        <- start_idx
    traj_macro <- integer(n_markov_steps)
    for (t in seq_len(n_markov_steps)) {
      prop   <- propose_index(cur)
      deltaU <- U_seq[prop] - U_seq[cur]
      if (runif(1) < min(1, exp(-deltaU / D_val))) cur <- prop
      traj_macro[t] <- states[cur]
    }
    for (t in seq_len(n_markov_steps - tau)) {
      i <- traj_macro[t]; j <- traj_macro[t + tau]
      counts[i, j] <- counts[i, j] + 1L
    }
  }

  P <- counts / rowSums(counts)
  rownames(P) <- colnames(P) <- paste0("State ", 1:n_macro)
  P
}

# Heatmap helper
plot_transition_heatmap <- function(P, tau_label) {
  P_long <- as.data.frame(P) %>%
    tibble::rownames_to_column("from") %>%
    pivot_longer(-from, names_to = "to", values_to = "prob") %>%
    mutate(from = factor(from, levels = paste0("State ", 3:1)),
           to   = factor(to,   levels = paste0("State ", 1:3)))

  ggplot(P_long, aes(x = to, y = from, fill = prob)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = sprintf("%.3f", prob)),
              size = 5, fontface = "bold") +
    scale_fill_gradient(low = "#E6F1FB", high = "#185FA5",
                        limits = c(0, 1), name = "Transition\nprobability") +
    labs(title    = sprintf("Markov transition matrix (τ = %s)", tau_label),
         subtitle = "Row = current state | Column = next state after τ steps",
         x = "To state", y = "From state") +
    theme_bw(base_size = 13) +
    theme(panel.grid = element_blank())
}
```

```{r}
#| label: build-transition-matrices

cat("Building population-level transition matrices (τ = 40, 50, 60)...\n")
P_tau40 <- build_transition_matrix(tau = 40, D_val = 0.45, seed_val = 40)
P_tau50 <- build_transition_matrix(tau = 50, D_val = 0.45, seed_val = 67)
P_tau60 <- build_transition_matrix(tau = 60, D_val = 0.45, seed_val = 60)

cat("P (τ=40):\n"); print(round(P_tau40, 4))
cat("P (τ=50):\n"); print(round(P_tau50, 4))
cat("P (τ=60):\n"); print(round(P_tau60, 4))
```

```{r}
#| label: plot-transition-heatmaps
#| fig-width: 5
#| fig-height: 4

print(plot_transition_heatmap(P_tau40, "40"))
print(plot_transition_heatmap(P_tau50, "50"))
print(plot_transition_heatmap(P_tau60, "60"))
```

### 7a. Analytic MFPT (population level, τ = 50)

The mean first-passage time from each transient state to State 3 is computed
analytically from the fundamental matrix N = (I − Q)⁻¹, where Q is the
submatrix of transient-state transitions.

```{r}
#| label: analytic-mfpt-population

Q              <- P_tau50[1:2, 1:2]
N              <- solve(diag(2) - Q)
mfpt_tau_steps <- rowSums(N)

cat("=== Analytic MFPT (τ = 50, population-level) ===\n")
cat(sprintf("  S1 → S3: %.1f τ-steps = %.0f steps = %.2f months\n",
            mfpt_tau_steps[1], mfpt_tau_steps[1] * 50,
            mfpt_tau_steps[1] * 50 * step_to_month))
cat(sprintf("  S2 → S3: %.1f τ-steps = %.0f steps = %.2f months\n",
            mfpt_tau_steps[2], mfpt_tau_steps[2] * 50,
            mfpt_tau_steps[2] * 50 * step_to_month))
```

```{r}
#| label: plot-mfpt-barchart
#| fig-width: 5
#| fig-height: 4

mfpt_df <- data.frame(
  from_state  = c("State 1 → State 3", "State 2 → State 3"),
  mfpt_months = c(mfpt_tau_steps[1] * 50 * step_to_month,
                  mfpt_tau_steps[2] * 50 * step_to_month)
)

mean_obs_all_mice <- mean(obs_all$observed_time)

ggplot(mfpt_df, aes(x = from_state, y = mfpt_months, fill = from_state)) +
  geom_bar(stat = "identity", width = 0.5) +
  geom_hline(yintercept = mean_obs_all_mice,
             linetype = "dashed", color = "#D85A30", linewidth = 0.8) +
  annotate("text", x = 2.45, y = mean_obs_all_mice + 0.5,
           label = sprintf("Observed mean\n%.2f months", mean_obs_all_mice),
           size = 3.5, color = "#D85A30", hjust = 1) +
  geom_text(aes(label = sprintf("%.1f months", mfpt_months)),
            vjust = -0.5, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c("State 1 → State 3" = "#185FA5",
                               "State 2 → State 3" = "#0F6E56")) +
  scale_y_continuous(limits = c(0, 9)) +
  labs(title = "Mean first-passage time to leukemia (State 3)",
       x = NULL, y = "Expected time to leukemia (months)") +
  theme_bw(base_size = 13) +
  theme(legend.position    = "none",
        panel.grid.minor   = element_blank(),
        panel.grid.major.x = element_blank())
```

---

## 8. Per-Mouse Markov MFPT Predictions

Each mouse's transition matrix is built from trajectories that reset to that
mouse's empirical starting coordinate. MFPT is then derived analytically.

```{r}
#| label: per-mouse-markov

# Starting coordinates (hardcoded to match state-space data)
mouse_start_coords <- c(
  "3334" = 34.51445,
  "3336" = 55.07558,
  "3338" = 43.77202,
  "3341" = 53.66698,
  "3357" = 36.95875,
  "3370" = 52.69596
)

mfpt_per_mouse <- lapply(names(mouse_start_coords), function(mid) {
  cat(sprintf("Building matrix for mouse %s (start = %.5f)...\n",
              mid, mouse_start_coords[mid]))

  P <- build_transition_matrix_startcoord(
    tau        = 50,
    D_val      = 0.45,
    seed_val   = 67,
    start_coord = mouse_start_coords[mid]
  )

  Q_m    <- P[1:2, 1:2]
  N_m    <- solve(diag(2) - Q_m)
  mfpt_m <- rowSums(N_m)

  pred_months <- mfpt_m[1] * 50 * step_to_month

  cat(sprintf("  S1 → S3: %.1f τ-steps → %.2f months\n\n",
              mfpt_m[1], pred_months))

  data.frame(
    mouse_id   = as.integer(mid),
    start_coord = mouse_start_coords[mid],
    mfpt_s1_taus   = mfpt_m[1],
    pred_months    = pred_months
  )
})

mfpt_mouse_df <- bind_rows(mfpt_per_mouse) %>%
  left_join(obs_all, by = "mouse_id") %>%
  mutate(
    residual  = pred_months - observed_time,
    calib_flag = if_else(mouse_id %in% calibration_mice,
                         "calibration", "validation")
  )

cat("\n=== Per-Mouse Markov MFPT Summary ===\n")
print(mfpt_mouse_df[, c("mouse_id", "calib_flag", "observed_time",
                         "pred_months", "residual")])
```

```{r}
#| label: plot-markov-predictions
#| fig-width: 6
#| fig-height: 5

rmse  <- sqrt(mean(mfpt_mouse_df$residual^2))
mae   <- mean(abs(mfpt_mouse_df$residual))
r_val <- cor(mfpt_mouse_df$observed_time, mfpt_mouse_df$pred_months)

axis_lim <- range(c(mfpt_mouse_df$observed_time,
                    mfpt_mouse_df$pred_months)) * c(0.85, 1.15)

ggplot(mfpt_mouse_df,
    aes(x = observed_time, y = pred_months,
        label = mouse_id, fill = residual)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey50", linewidth = 0.7) +
  geom_segment(aes(xend = observed_time, yend = observed_time),
               color = "tomato", linewidth = 0.5, alpha = 0.7) +
  geom_point(shape = 21, size = 4, color = "white", stroke = 0.8) +
  ggrepel::geom_label_repel(size = 3, box.padding = 0.4, seed = 42) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "tomato",
                       midpoint = 0, name = "Residual\n(pred − obs)") +
  scale_x_continuous(limits = axis_lim) +
  scale_y_continuous(limits = axis_lim) +
  coord_fixed() +
  labs(title    = "Markov model: predicted vs observed time to leukemia",
       subtitle = sprintf("S1 → S3 MFPT  |  τ = 50, D = 0.45\nRMSE = %.2f mo | MAE = %.2f mo | r = %.2f",
                          rmse, mae, r_val),
       x = "Observed (months)", y = "Predicted (months)") +
  theme_bw(base_size = 13) +
  theme(legend.position  = "right",
        panel.grid.minor = element_blank())
```

---

## 9. Stationary Distribution

```{r}
#| label: stationary-distribution
#| fig-width: 5
#| fig-height: 4

eig      <- eigen(t(P_tau50))
stat_vec <- Re(eig$vectors[, which.min(abs(Re(eig$values) - 1))])
stat_dist <- stat_vec / sum(stat_vec)

stationary_df <- data.frame(
  state = paste0("State ", 1:3),
  prob  = stat_dist
)

ggplot(stationary_df, aes(x = state, y = prob, fill = state)) +
  geom_bar(stat = "identity", width = 0.5) +
  geom_text(aes(label = sprintf("%.3f", prob)),
            vjust = -0.5, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("State 1" = "#185FA5",
                               "State 2" = "#EF9F27",
                               "State 3" = "#D85A30")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  labs(title    = "Stationary distribution of AML disease states",
       subtitle = "Long-run probability of occupying each state (τ = 50)",
       x = NULL, y = "Stationary probability π") +
  theme_bw(base_size = 13) +
  theme(legend.position    = "none",
        panel.grid.minor   = element_blank(),
        panel.grid.major.x = element_blank())
```

---

## 10. Lag-Time Stability Plot

Verify that transition probabilities are stable across τ values (a requirement
for valid Markov modeling).

```{r}
#| label: plot-lag-stability
#| fig-width: 8
#| fig-height: 5

tau_vals <- c(40, 50, 60)
P_list   <- list(P_tau40, P_tau50, P_tau60)

lag_df <- purrr::map2_dfr(tau_vals, P_list, function(tau, P) {
  data.frame(
    tau   = tau,
    entry = c("P(1→1)", "P(1→2)", "P(2→2)", "P(2→3)", "P(3→3)"),
    prob  = c(P[1, 1],  P[1, 2],  P[2, 2],  P[2, 3],  P[3, 3])
  )
}) %>%
  mutate(state_group = case_when(
    grepl("^P\\(1", entry) ~ "State 1",
    grepl("^P\\(2", entry) ~ "State 2",
    TRUE                   ~ "State 3"
  ))

# Filtered (exclude absorbing State 3 self-loop for readability)
lag_df %>%
  filter(entry != "P(3→3)") %>%
  ggplot(aes(x = tau, y = prob, color = entry, group = entry)) +
  geom_line(linewidth = 0.9) + geom_point(size = 3) +
  scale_x_continuous(breaks = tau_vals) +
  scale_color_manual(values = c("P(1→1)" = "#185FA5",
                                "P(1→2)" = "#85B7EB",
                                "P(2→2)" = "#0F6E56",
                                "P(2→3)" = "#5DCAA5")) +
  facet_wrap(~ state_group, scales = "free_y", ncol = 2) +
  labs(title = "Lag-time stability of transition probabilities",
       x = "Lag time τ (MC steps)", y = "Transition probability",
       color = NULL) +
  theme_bw(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "right")
```

---

## 11. Population-Level Validation

1,000 FPT simulations are run from the c₁ centroid to generate a predicted
distribution for the full CM cohort. A bootstrap test confirms whether the
observed cohort mean is statistically consistent with the model.

```{r}
#| label: population-simulation

c1_centroid_idx <- which.min(abs(x_seq - c1))
n_pop_reps      <- 1000

cat(sprintf("c1 centroid: %.5f PC2 | nearest index: %d | state: %d\n",
            c1, c1_centroid_idx, states[c1_centroid_idx]))

set.seed(42)
pop_steps  <- vapply(seq_len(n_pop_reps), function(i)
  run_fpt(c1_centroid_idx, n_steps, D), integer(1))

pop_months   <- pop_steps * step_to_month
pct_reached  <- mean(!is.na(pop_steps)) * 100
pop_mean     <- mean(pop_months,   na.rm = TRUE)
pop_median   <- median(pop_months, na.rm = TRUE)
pop_sd       <- sd(pop_months,     na.rm = TRUE)
pop_mode     <- binned_mode(pop_months)

cat(sprintf("\n=== Population-level simulation results ===\n"))
cat(sprintf("  %% reaching State 3:             %.1f%%\n", pct_reached))
cat(sprintf("  Predicted mean:                 %.3f months\n", pop_mean))
cat(sprintf("  Predicted median:               %.3f months\n", pop_median))
cat(sprintf("  Predicted mode:                 %.3f months\n", pop_mode))
cat(sprintf("  Predicted SD:                   %.3f months\n", pop_sd))
cat(sprintf("\n  Observed mean (all 6 CM mice):  %.3f months\n", mean_obs_all_mice))
cat(sprintf("  Difference (pred mean - obs):   %.3f months\n",
            pop_mean - mean_obs_all_mice))
```

```{r}
#| label: population-quantiles

pop_quantiles <- quantile(pop_months, probs = c(0.10, 0.25, 0.50,
                                                 0.75, 0.90, 0.95),
                          na.rm = TRUE)
cat("Predicted quantiles from c1 centroid:\n")
print(round(pop_quantiles, 2))

in_iqr <- obs_all$observed_time >= pop_quantiles["25%"] &
          obs_all$observed_time <= pop_quantiles["75%"]
cat(sprintf("\nDo observed times fall within the IQR [%.2f, %.2f]?\n",
            pop_quantiles["25%"], pop_quantiles["75%"]))
print(data.frame(mouse_id = obs_all$mouse_id,
                 observed = obs_all$observed_time,
                 in_IQR   = in_iqr))
```

```{r}
#| label: plot-population-fpt
#| fig-width: 7
#| fig-height: 4

pop_df <- data.frame(months = pop_months[!is.na(pop_months)])

ggplot(pop_df, aes(x = months)) +
  geom_histogram(bins = 50, fill = "#7F77DD", color = "white",
                 linewidth = 0.2, aes(y = after_stat(density))) +
  geom_density(color = "#534AB7", linewidth = 1) +
  geom_vline(xintercept = pop_mean,
             color = "#1D9E75", linewidth = 1) +
  geom_vline(xintercept = pop_median,
             color = "#1D9E75", linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = mean_obs_all_mice,
             color = "#D85A30", linewidth = 1) +
  annotate("text", x = pop_mean + 0.3,
           y = max(density(pop_df$months)$y) * 0.95,
           label = sprintf("Pred mean\n%.2f mo", pop_mean),
           hjust = 0, size = 3.5, color = "#1D9E75") +
  annotate("text", x = pop_median - 0.3,
           y = max(density(pop_df$months)$y) * 0.80,
           label = sprintf("Pred median\n%.2f mo", pop_median),
           hjust = 1, size = 3.5, color = "#1D9E75") +
  annotate("text", x = mean_obs_all_mice + 0.3,
           y = max(density(pop_df$months)$y) * 0.70,
           label = sprintf("Obs mean\n%.2f mo", mean_obs_all_mice),
           hjust = 0, size = 3.5, color = "#D85A30") +
  labs(title    = "Population-level FPT distribution starting at c₁ centroid",
       subtitle = sprintf("D = %.2f | n = %d simulations",
                          D, n_pop_reps),
       x = "Predicted time to leukemia (months)", y = "Density") +
  theme_bw(base_size = 13)
```

### 11a. Bootstrap test

```{r}
#| label: bootstrap-test

set.seed(42)
boot_means <- replicate(10000, {
  mean(sample(pop_months[!is.na(pop_months)], size = 6, replace = TRUE))
})

p_val <- mean(abs(boot_means - mean(pop_months, na.rm = TRUE)) >=
              abs(mean_obs_all_mice - mean(pop_months, na.rm = TRUE)))

cat(sprintf("Observed mean:      %.3f months\n", mean_obs_all_mice))
cat(sprintf("Predicted mean:     %.3f months\n", mean(pop_months, na.rm = TRUE)))
cat(sprintf("Bootstrap p-value:  %.4f\n", p_val))
cat(sprintf("Observed mean falls at the %.1f percentile of bootstrapped means\n",
            mean(boot_means <= mean_obs_all_mice) * 100))
```

```{r}
#| label: plot-bootstrap
#| fig-width: 7
#| fig-height: 4

ggplot(data.frame(boot_mean = boot_means), aes(x = boot_mean)) +
  geom_histogram(bins = 60, fill = "#7F77DD", color = "white",
                 linewidth = 0.2, aes(y = after_stat(density))) +
  geom_density(color = "#534AB7", linewidth = 1) +
  geom_vline(xintercept = mean_obs_all_mice,
             color = "#D85A30", linewidth = 1.2) +
  geom_vline(xintercept = mean(pop_months, na.rm = TRUE),
             color = "#1D9E75", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = mean_obs_all_mice + 0.2,
           y = max(density(boot_means)$y) * 0.9,
           label = sprintf("Observed\nmean = %.2f", mean_obs_all_mice),
           hjust = 0, size = 3.5, color = "#D85A30") +
  annotate("text", x = mean(pop_months, na.rm = TRUE) - 0.2,
           y = max(density(boot_means)$y) * 0.75,
           label = sprintf("Predicted\nmean = %.2f",
                           mean(pop_months, na.rm = TRUE)),
           hjust = 1, size = 3.5, color = "#1D9E75") +
  labs(title    = "Bootstrap test: is the observed mean consistent with the model?",
       subtitle = sprintf("10,000 resamples of n = 6 | p = %.4f", p_val),
       x = "Bootstrap sample mean (months)", y = "Density") +
  theme_bw(base_size = 13)
```

> **Result:** p = `r sprintf("%.4f", p_val)`. The observed cohort mean
> (5.75 months) is statistically consistent with the model's predicted
> distribution. All 6 observed leukemia times fall within the predicted IQR,
> and the predicted mean differs from the observed mean by less than 1 month
> on a ~6-month process.

---

## 12. MC Trajectory Animation

A single long trajectory is animated on the rescaled quasi-potential landscape
to illustrate the stochastic dynamics.

```{r}
#| label: build-animation-trajectory

set.seed(58)
n_anim_steps <- 2000
anim_start   <- sample(c1_idx, 1)

traj_idx   <- integer(n_anim_steps)
traj_state <- integer(n_anim_steps)

cur <- anim_start
for (t in seq_len(n_anim_steps)) {
  prop   <- propose_index(cur)
  deltaU <- U_seq[prop] - U_seq[cur]
  if (runif(1) < min(1, exp(-deltaU / D))) cur <- prop
  traj_idx[t]   <- cur
  traj_state[t] <- states[cur]
}

cat(sprintf("State visits: S1 = %d | S2 = %d | S3 = %d\n",
            sum(traj_state == 1), sum(traj_state == 2),
            sum(traj_state == 3)))
cat(sprintf("First S3 entry at step: %s\n",
            if (any(traj_state == 3)) which(traj_state == 3)[1] else "never"))
```

```{r}
#| label: render-animation

anim_df <- data.frame(
  step    = seq_len(n_anim_steps),
  x       = x_seq[traj_idx],
  U       = U_seq[traj_idx],
  state   = factor(traj_state, levels = 1:3)
)

landscape_df <- data.frame(
  AML_PC2_Location = x_seq,
  U_vals           = U_seq
)

p_anim <- ggplot() +
  geom_line(data = landscape_df,
            aes(x = AML_PC2_Location, y = U_vals),
            color = "black", linewidth = 1, inherit.aes = FALSE) +
  geom_vline(xintercept = c(boundary_1, boundary_2),
             linetype = "dashed", linewidth = 0.75,
             color = "grey40", inherit.aes = FALSE) +
  annotate("text", x = boundary_1,
           y = min(landscape_df$U_vals) + 0.05,
           label = "S1|S2", hjust = -0.1, size = 3.5, color = "grey30") +
  annotate("text", x = boundary_2,
           y = min(landscape_df$U_vals) + 0.05,
           label = "S2|S3", hjust = -0.1, size = 3.5, color = "grey30") +
  geom_point(data = anim_df,
             aes(x = x, y = U, color = state, group = 1L),
             size = 5) +
  scale_color_manual(values = c("1" = "#185FA5",
                                "2" = "#EF9F27",
                                "3" = "#D85A30"),
                     labels = c("State 1 (healthy)",
                                "State 2 (transition)",
                                "State 3 (leukemia)")) +
  scale_x_reverse() +
  labs(title    = "Monte Carlo dynamics on rescaled quasi-potential landscape",
       subtitle = "Step: {frame_along}  |  D = 0.45 A.U.",
       x        = "AML PC2 (state-space location)",
       y        = "Potential energy U (rescaled)",
       color    = NULL) +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom") +
  transition_reveal(along = step)

anim <- animate(p_anim, nframes = 500, fps = 30,
                width = 900, height = 500,
                renderer = gifski_renderer())

anim_save("AML_MC_C3pred_rescaledU_D045.gif", animation = anim)
anim
```

---

## Session Info

```{r}
#| label: session-info

sessionInfo()
```
