---
title: "AML TradeSeq Trajectory Analysis and Pathway Enrichment"
format: html
author: Danya Hassan
knitr:
  opts_chunk:
    echo: TRUE
    warning: FALSE
    error: TRUE
---

## Overview

This document fits gene expression trajectories along the AML pseudotime axis
using **tradeSeq**, clusters genes by their CM–Ctrl divergence pattern, and
links divergence clusters to biological pathways via ORA (clusterProfiler) and
GSEA. The analysis is restricted to the top 250 PC2 eigengenes identified in
`aml2018_state_space_construction.qmd`.

Pipeline summary:

1. Filter samples and extract raw counts
2. Map Ensembl IDs → gene symbols (biomaRt)
3. Subset to top 250 eigengenes by PC2 loading
4. Fit tradeSeq GAMs along pseudotime (Ctrl and CM lineages)
5. Cluster genes by divergence trajectory (hierarchical, k = 6)
6. Visualise smooth trajectories per cluster
7. Over-representation analysis (ORA) per cluster
8. GSEA using LFC-ranked gene list
9. STRING network analysis (optional)

---

## 1. Setup

```{r}
#| label: libraries

library(tradeSeq)
library(SingleCellExperiment)
library(slingshot)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(patchwork)
library(readr)
library(stringr)
library(biomaRt)
library(clusterProfiler)
library(org.Mm.eg.db)
library(enrichplot)
library(STRINGdb)
```

```{r}
#| label: load-data

eigengenes  <- read_csv("eigengenes_PC2_aml2018.csv")
mrna_2018   <- readRDS("AML_mRNA_2018_1_real.rds")
aml_ss      <- read_csv("final_aml_mrna_state_space.csv")
all_2018_ss <- read_csv("AML_2018_StateSpace_Mice.csv")
```

---

## 2. State-Space Boundaries and Sample Filtering

```{r}
#| label: state-thresholds

state_thresholds <- c(
  c1_perturbation   =  14.89216,
  c2_energy_barrier = -40.56986,
  c3_leukemia       = -299.0184
)
```

```{r}
#| label: filter-samples

# Exclude outlier mice identified during state-space construction
lib_ids_keep <- all_2018_ss |>
  filter(
    mouse_id.x != 3368,
    !(mouse_id.x == 3336 & Time == 8.0),
    !(mouse_id.x == 3357 & Time >  3.0)
  ) |>
  pull(library_id) |>
  unique()

ctrl_lib_ids <- all_2018_ss |> filter(group_name == "Ctrl") |> pull(library_id) |> unique()
cm_lib_ids   <- all_2018_ss |> filter(group_name == "CM")   |> pull(library_id) |> unique()

stopifnot(length(lib_ids_keep) == length(ctrl_lib_ids) + length(cm_lib_ids))

cat(sprintf("Samples retained: %d (%d Ctrl, %d CM)\n",
            length(lib_ids_keep), length(ctrl_lib_ids), length(cm_lib_ids)))
```

```{r}
#| label: plot-state-space
#| fig-width: 8
#| fig-height: 5

state_bands <- tibble(
  ymin  = c(-Inf,    -250.0, -40.5699,  14.89163),
  ymax  = c(-250.0, -40.5699, 14.89163,  Inf),
  fill  = c("#FDE68A", "#D1D5DB", "#BFDBFE", "#F0FDF4"),
  label = c("State 3 (Leukemia)", "State 2 (Energy Barrier)",
            "State 1 (Perturbation)", "Normal")
)

ggplot(aml_ss, aes(x = Time, y = AML_space.calc,
                   colour = factor(mouse_id))) +
  geom_rect(data = state_bands, inherit.aes = FALSE,
            aes(xmin = -Inf, xmax = Inf, ymin = ymin, ymax = ymax,
                fill = label), alpha = 0.25) +
  scale_fill_manual(values = setNames(state_bands$fill, state_bands$label),
                    name = "State") +
  geom_point(size = 2) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = c(-250, -40.5699, 14.89163),
             linetype = "dashed", color = "red", linewidth = 0.7) +
  labs(title  = "AML and Ctrl Mice: State-Space Projection",
       x      = "Time (months)",
       y      = "AML State-Space Location",
       colour = "Mouse ID") +
  theme_bw(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5))
```

---

## 3. Extract and Filter Raw Counts

```{r}
#| label: extract-counts

mrna2018_counts_raw <- as.data.frame(assay(mrna_2018, "counts"))
clean_counts        <- mrna2018_counts_raw[,
  colnames(mrna2018_counts_raw) %in% lib_ids_keep]

cat(sprintf("Dimensions after sample filtering: %d genes × %d samples\n",
            nrow(clean_counts), ncol(clean_counts)))
```

---

## 4. Map Ensembl IDs to Gene Symbols

```{r}
#| label: gene-symbol-mapping

mart <- useMart(biomart = "ensembl", dataset = "mmusculus_gene_ensembl")

gene_annot <- getBM(
  attributes = c("ensembl_gene_id", "external_gene_name", "gene_biotype"),
  filters    = "ensembl_gene_id",
  values     = rownames(clean_counts),
  mart       = mart
)

# Named lookup: Ensembl ID → gene symbol
gene_map_vec <- setNames(gene_annot$external_gene_name,
                         gene_annot$ensembl_gene_id)

# Map; fall back to Ensembl ID where no symbol found
new_rownames <- gene_map_vec[rownames(clean_counts)]
new_rownames[is.na(new_rownames)] <- rownames(clean_counts)[is.na(new_rownames)]
new_rownames <- make.unique(new_rownames)
rownames(clean_counts) <- new_rownames

cat(sprintf("Gene symbol mapping: %d / %d rows annotated\n",
            sum(!is.na(gene_map_vec[rownames(clean_counts)])),
            nrow(clean_counts)))
```

---

## 5. Subset to Top 250 Eigengenes

Only genes with the highest absolute PC2 loading (variance-scaled) are retained.
These are the genes that most strongly drive transcriptomic movement along the
AML state-space axis.

```{r}
#| label: subset-eigengenes

top_250_genes <- eigengenes |>
  mutate(eigengene_PC2_scaled = as.numeric(eigengene_PC2_scaled)) |>
  arrange(desc(abs(eigengene_PC2_scaled))) |>
  slice_head(n = 250) |>
  pull(gene)

counts_eigengenes <- clean_counts[rownames(clean_counts) %in% top_250_genes, ]

cat(sprintf("Eigengene-filtered dimensions: %d genes × %d samples\n",
            nrow(counts_eigengenes), ncol(counts_eigengenes)))
```

---

## 6. Align Metadata

```{r}
#| label: align-metadata

meta <- all_2018_ss |>
  filter(library_id %in% lib_ids_keep) |>
  dplyr::slice(match(colnames(counts_eigengenes), library_id))

stopifnot(all(meta$library_id == colnames(counts_eigengenes)))
```

---

## 7. Build tradeSeq Inputs

Two lineages are modeled: **Ctrl** (healthy trajectory) and **CM** (leukemic
trajectory). Pseudotime is each sample's observed collection time in months.
The binary cell-weight matrix assigns each sample to exactly one lineage.

```{r}
#| label: tradseq-inputs

lineages   <- c("Ctrl", "CM")
n_cells    <- ncol(counts_eigengenes)
n_lineages <- length(lineages)

# Pseudotime matrix (NAs replaced with 0 per tradeSeq requirements)
pseudotime <- matrix(NA_real_, nrow = n_cells, ncol = n_lineages,
                     dimnames = list(meta$library_id, lineages))
pseudotime[meta$group_name == "Ctrl", "Ctrl"] <- meta$Time[meta$group_name == "Ctrl"]
pseudotime[meta$group_name == "CM",   "CM"]   <- meta$Time[meta$group_name == "CM"]
pseudotime[is.na(pseudotime)] <- 0

# Binary cell-weight matrix
cellWeights <- matrix(0, nrow = n_cells, ncol = n_lineages,
                      dimnames = list(meta$library_id, lineages))
cellWeights[meta$group_name == "Ctrl", "Ctrl"] <- 1
cellWeights[meta$group_name == "CM",   "CM"]   <- 1

stopifnot(identical(rownames(pseudotime), rownames(cellWeights)),
          identical(rownames(pseudotime), colnames(counts_eigengenes)))
```

---

## 8. Select Number of Knots and Fit GAMs

```{r}
#| label: evaluate-k

set.seed(7)
icMat <- evaluateK(
  counts      = as.matrix(counts_eigengenes),
  pseudotime  = pseudotime,
  cellWeights = cellWeights,
  k           = 3:9,
  nGenes      = nrow(counts_eigengenes),
  verbose     = TRUE
)
```

```{r}
#| label: fit-gam

# k = 8 selected based on AIC/BIC elbow from icMat
set.seed(9)
sce <- fitGAM(
  counts      = as.matrix(counts_eigengenes),
  pseudotime  = pseudotime,
  cellWeights = cellWeights,
  nknots      = 8,
  verbose     = FALSE
)

cat("GAM convergence:\n")
print(table(rowData(sce)$tradeSeq$converged))
```

---

## 9. Predict Smooth Trajectories

```{r}
#| label: predict-smooth

nPointsClus <- 20

yhat <- predictSmooth(
  sce,
  gene    = rownames(counts_eigengenes),
  nPoints = nPointsClus
) |>
  mutate(lineage = if_else(lineage == 1, "Ctrl", "CM"))

cat(sprintf("Smooth predictions: %d rows\n", nrow(yhat)))
head(yhat)
```

---

## 10. Divergence-Based Gene Clustering

Genes are clustered by their CM − Ctrl divergence trajectory (hierarchical
clustering on scaled expression, k = 6). Clustering on divergence shape rather
than magnitude makes the clusters interpretable as distinct biological programs.

```{r}
#| label: divergence-clustering

# Build wide matrix: genes × (Ctrl_t1…Ctrl_t20 | CM_t1…CM_t20)
yhat_wide <- yhat |>
  group_by(gene, lineage) |>
  mutate(time_id = paste0(lineage, "_t", row_number())) |>
  ungroup() |>
  select(gene, time_id, yhat) |>
  pivot_wider(names_from = time_id, values_from = yhat)

yhat_mat <- as.matrix(yhat_wide[, -1])
rownames(yhat_mat) <- yhat_wide$gene

# Scale within each gene (cluster on shape, not magnitude)
yhat_scaled <- t(scale(t(yhat_mat)))

# Compute CM − Ctrl divergence per gene
ctrl_cols <- grep("^Ctrl_", colnames(yhat_scaled))
cm_cols   <- grep("^CM_",   colnames(yhat_scaled))
delta_mat <- yhat_scaled[, cm_cols] - yhat_scaled[, ctrl_cols]

# Hierarchical clustering on divergence
set.seed(42)
k             <- 6
hc_delta      <- hclust(dist(delta_mat), method = "ward.D2")
clusterLabels <- cutree(hc_delta, k = k)

cluster_df <- tibble(
  gene        = names(clusterLabels),
  div_cluster = factor(clusterLabels)
)

cat("Genes per divergence cluster:\n")
print(table(cluster_df$div_cluster))
```

```{r}
#| label: build-scaled-long

# Long-format scaled expression with actual time values
yhat_scaled_long <- yhat_mat |>
  as.data.frame() |>
  rownames_to_column("gene") |>
  pivot_longer(-gene, names_to = "time_id", values_to = "yhat_scaled") |>
  mutate(
    lineage  = if_else(str_starts(time_id, "Ctrl"), "Ctrl", "CM"),
    time_idx = as.integer(str_extract(time_id, "\\d+"))
  )

time_lookup <- yhat |>
  group_by(lineage) |>
  mutate(time_idx = row_number()) |>
  ungroup() |>
  select(lineage, time_idx, time) |>
  distinct()

plot_df_scaled <- yhat_scaled_long |>
  left_join(time_lookup, by = c("lineage", "time_idx")) |>
  inner_join(cluster_df, by = "gene")
```

---

## 11. Trajectory Plots

```{r}
#| label: trajectory-plot-functions

lineage_colors <- c("Ctrl" = "orange", "CM" = "darkseagreen3")

# Spaghetti + mean overlay
plot_cluster_full <- function(cl, df = plot_df_scaled) {
  ggplot(df |> filter(div_cluster == cl),
         aes(x = time, y = yhat_scaled,
             color = lineage,
             group = interaction(gene, lineage))) +
    geom_line(alpha = 0.3, linewidth = 0.7) +
    stat_summary(aes(group = lineage), fun = mean, geom = "line",
                 linewidth = 1.8) +
    scale_color_manual(values = lineage_colors) +
    labs(title = paste("Divergence Cluster", cl),
         x = "Pseudotime", y = "Scaled expression",
         color = "Lineage") +
    theme_classic(base_size = 13) +
    theme(plot.title = element_text(hjust = 0.5))
}

# Mean ± SE only
plot_cluster_mean <- function(cl, df = plot_df_scaled) {
  ggplot(df |> filter(div_cluster == cl),
         aes(x = time, y = yhat_scaled, color = lineage)) +
    stat_summary(fun = mean, geom = "line", linewidth = 1.8) +
    stat_summary(fun.data = mean_se, geom = "ribbon",
                 aes(fill = lineage, group = lineage),
                 alpha = 0.15, color = NA) +
    scale_color_manual(values = lineage_colors) +
    scale_fill_manual(values  = lineage_colors) +
    labs(title = paste("Cluster", cl, "— Mean ± SE"),
         x = "Pseudotime", y = "Mean scaled expression",
         color = "Lineage") +
    theme_classic(base_size = 13) +
    theme(plot.title = element_text(hjust = 0.5))
}
```

### 11a. All clusters — mean trajectory overview

```{r}
#| label: plot-all-clusters-mean
#| fig-width: 10
#| fig-height: 7

ggplot(plot_df_scaled, aes(x = time, y = yhat_scaled, color = lineage)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.6) +
  stat_summary(fun.data = mean_se, geom = "ribbon",
               aes(fill = lineage, group = lineage),
               alpha = 0.15, color = NA) +
  facet_wrap(~ div_cluster, ncol = 3, scales = "free_y") +
  scale_color_manual(values = lineage_colors) +
  scale_fill_manual(values  = lineage_colors) +
  labs(title = "Gene Clusters: CM vs Ctrl Divergence (Mean ± SE)",
       x = "Pseudotime", y = "Mean scaled expression",
       color = "Lineage", fill = "Lineage") +
  theme_classic(base_size = 13) +
  theme(strip.text = element_text(face = "bold"),
        plot.title = element_text(hjust = 0.5))
```

### 11b. Per-cluster spaghetti + mean

```{r}
#| label: plot-per-cluster
#| fig-width: 10
#| fig-height: 4

lapply(levels(plot_df_scaled$div_cluster), function(cl) {
  plot_cluster_full(cl) | plot_cluster_mean(cl)
})
```

---

## 12. Gene Universe Construction for Enrichment

Two background gene sets are constructed from the full mrna_2018 object:

- **expressed_universe** — all genes with mean TPM > 1 across retained samples
- **variable_universe** — expressed genes in the top 50% of coefficient of
  variation (removes housekeeping genes; used for both ORA and GSEA)

```{r}
#| label: build-gene-universe

# --- TPM-based expressed universe ---
tpm_mat   <- as.data.frame(assay(mrna_2018, "tpm"))
tpm_clean <- tpm_mat[, colnames(tpm_mat) %in% lib_ids_keep]

full_universe_names <- gene_map_vec[rownames(tpm_clean)]
full_universe_names[is.na(full_universe_names)] <-
  rownames(tpm_clean)[is.na(full_universe_names)]
full_universe_names <- make.unique(full_universe_names)

tpm_means         <- rowMeans(tpm_clean)
expressed_universe <- full_universe_names[tpm_means > 1]

cat(sprintf("Expressed universe: %d genes\n", length(expressed_universe)))

# --- Variance-filtered universe (top 50% CV among expressed genes) ---
counts_scaled_mat   <- as.data.frame(assay(mrna_2018, "counts_scaled"))
counts_scaled_clean <- counts_scaled_mat[,
  colnames(counts_scaled_mat) %in% lib_ids_keep]

scaled_names <- gene_map_vec[rownames(counts_scaled_clean)]
scaled_names[is.na(scaled_names)] <-
  rownames(counts_scaled_clean)[is.na(scaled_names)]
scaled_names <- make.unique(scaled_names)
rownames(counts_scaled_clean) <- scaled_names

gene_means <- rowMeans(counts_scaled_clean)
gene_cv    <- apply(counts_scaled_clean, 1, sd) / (gene_means + 1)
min_expr   <- gene_means > 1
cv_thresh  <- quantile(gene_cv[min_expr], 0.5)

variable_universe <- names(gene_cv[min_expr & gene_cv >= cv_thresh])
cat(sprintf("Variable universe: %d genes\n", length(variable_universe)))
```

```{r}
#| label: universe-cluster-overlap

cat("Cluster gene overlap with variable_universe:\n")
for (cl in levels(cluster_df$div_cluster)) {
  n <- length(intersect(cluster_df$gene[cluster_df$div_cluster == cl],
                        variable_universe))
  cat(sprintf("  Cluster %s: %d genes\n", cl, n))
}
```

---

## 13. Over-Representation Analysis (ORA) per Cluster

ORA asks whether GO Biological Process terms are statistically enriched in each
cluster's gene list relative to the variable universe background.

```{r}
#| label: run-ora

ora_results_full <- lapply(levels(cluster_df$div_cluster), function(cl) {
  genes_in_cluster <- intersect(
    cluster_df$gene[cluster_df$div_cluster == cl],
    expressed_universe
  )

  enrichGO(
    gene          = genes_in_cluster,
    universe      = variable_universe,
    OrgDb         = org.Mm.eg.db,
    keyType       = "SYMBOL",
    ont           = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.2,
    readable      = TRUE
  )
})
names(ora_results_full) <- paste0("Cluster_", levels(cluster_df$div_cluster))

cat("Significant GO terms per cluster:\n")
for (cl in names(ora_results_full)) {
  n <- nrow(as.data.frame(ora_results_full[[cl]]))
  cat(sprintf("  %s: %d terms\n", cl, n))
}
```

### 13a. Top terms per cluster — dot plot

```{r}
#| label: plot-ora-top
#| fig-width: 9
#| fig-height: 6

top_n_terms <- 5

ora_top_df <- bind_rows(lapply(names(ora_results_full), function(cl) {
  df <- as.data.frame(ora_results_full[[cl]])
  if (nrow(df) == 0) return(NULL)
  df |>
    arrange(p.adjust) |>
    slice_head(n = top_n_terms) |>
    mutate(Cluster = cl)
}))

if (nrow(ora_top_df) > 0) {
  ggplot(ora_top_df, aes(x = Cluster, y = reorder(Description, -p.adjust),
                          size = Count, color = p.adjust)) +
    geom_point() +
    scale_color_continuous(trans = "reverse") +
    labs(x = "Cluster", y = "GO Term",
         color = "Adj. p-value", size = "Gene count") +
    theme_minimal(base_size = 12) +
    theme(axis.text.y = element_text(size = 9.5))
} else {
  cat("No significant GO terms to plot.\n")
}
```

### 13b. Curated biological terms

Selected terms of direct relevance to AML disease biology:

```{r}
#| label: plot-ora-curated
#| fig-width: 8
#| fig-height: 5

selected_terms <- list(
  Cluster_1 = c("positive regulation of exocytosis",
                "positive regulation of mast cell degranulation",
                "mast cell degranulation"),
  Cluster_2 = c("positive regulation of angiogenesis",
                "endothelial cell proliferation"),
  Cluster_4 = c("heterochromatin formation",
                "epigenetic regulation of gene expression",
                "chromatin remodeling"),
  Cluster_6 = c("regulation of transcription by RNA polymerase I",
                "cytoplasmic translational initiation")
)

ora_curated_df <- bind_rows(lapply(names(selected_terms), function(cl) {
  df <- as.data.frame(ora_results_full[[cl]])
  if (nrow(df) == 0) { cat(cl, ": no results\n"); return(NULL) }
  matched <- df |> filter(Description %in% selected_terms[[cl]])
  cat(sprintf("%s: matched %d / %d curated terms\n",
              cl, nrow(matched), length(selected_terms[[cl]])))
  matched$Cluster <- cl
  matched
}))

if (nrow(ora_curated_df) > 0) {
  ggplot(ora_curated_df,
         aes(x = Cluster,
             y = reorder(Description, -p.adjust),
             size = Count, color = p.adjust)) +
    geom_point() +
    scale_color_continuous(trans = "reverse") +
    labs(title  = "Curated GO terms by divergence cluster",
         x      = "Cluster",
         y      = NULL,
         color  = "Adj. p-value",
         size   = "Gene count") +
    theme_minimal(base_size = 12) +
    theme(axis.text.y = element_text(size = 9.5))
} else {
  cat("No curated terms matched.\n")
}
```

### 13c. Cluster 2 — angiogenesis gene trajectories

```{r}
#| label: plot-angiogenesis-genes
#| fig-width: 7
#| fig-height: 4

cluster2_angio_genes <- ora_results_full$Cluster_2@result |>
  filter(Description == "positive regulation of angiogenesis") |>
  pull(geneID) |>
  strsplit("/") |>
  unlist()

plot_df_scaled |>
  filter(gene %in% cluster2_angio_genes, div_cluster == 2) |>
  ggplot(aes(x = time, y = yhat_scaled,
             color = lineage,
             group = interaction(gene, lineage))) +
  geom_line(alpha = 0.4) +
  stat_summary(aes(group = lineage), fun = mean,
               geom = "line", linewidth = 1.8) +
  scale_color_manual(values = lineage_colors) +
  labs(title = "Angiogenesis genes — Cluster 2 trajectory",
       x = "Pseudotime", y = "Scaled expression",
       color = "Lineage") +
  theme_classic(base_size = 13)
```

---

## 14. GSEA — Global LFC Ranking

GSEA uses a continuous CM vs Ctrl log₂ fold-change ranking across the entire
variable universe, providing more statistical power than the binary ORA.

```{r}
#| label: build-gsea-ranking

cm_samples   <- meta$library_id[meta$group_name == "CM"]
ctrl_samples <- meta$library_id[meta$group_name == "Ctrl"]

mean_cm   <- rowMeans(counts_scaled_clean[, cm_samples,   drop = FALSE])
mean_ctrl <- rowMeans(counts_scaled_clean[, ctrl_samples, drop = FALSE])

lfc <- log2((mean_cm + 1) / (mean_ctrl + 1))

gsea_ranked <- sort(lfc[names(lfc) %in% variable_universe], decreasing = TRUE)
gsea_ranked <- gsea_ranked[!is.na(gsea_ranked) & is.finite(gsea_ranked)]

cat(sprintf("Genes in GSEA ranked list: %d\n", length(gsea_ranked)))
```

```{r}
#| label: run-gsea-go

gsea_go_full <- gseGO(
  geneList     = gsea_ranked,
  OrgDb        = org.Mm.eg.db,
  keyType      = "SYMBOL",
  ont          = "BP",
  minGSSize    = 15,
  maxGSSize    = 500,
  pvalueCutoff = 0.05,
  eps          = 0,
  verbose      = FALSE
)

cat(sprintf("Significant GSEA GO terms: %d\n",
            nrow(as.data.frame(gsea_go_full))))
```

```{r}
#| label: gsea-cluster-membership

# Link GSEA core-enrichment genes back to divergence clusters
gsea_sig <- as.data.frame(gsea_go_full) |>
  filter(p.adjust < 0.05) |>
  arrange(p.adjust)

gsea_sig$cluster_membership <- sapply(gsea_sig$core_enrichment, function(gene_str) {
  genes   <- strsplit(gene_str, "/")[[1]]
  cl_hits <- cluster_df |>
    filter(gene %in% genes) |>
    count(div_cluster, sort = TRUE)
  paste(paste0("C", cl_hits$div_cluster, "(n=", cl_hits$n, ")"), collapse = ", ")
})

gsea_sig |>
  select(Description, p.adjust, NES, cluster_membership) |>
  head(20)
```

```{r}
#| label: plot-gsea-enrichment
#| fig-width: 9
#| fig-height: 5

gseaplot2(gsea_go_full,
          geneSetID = 1:5,
          title     = "Top GSEA GO Terms (CM vs Ctrl LFC ranking)")
```

---

## 15. STRING Network Analysis (Optional)

For clusters with interpretable GO enrichment, STRING protein-interaction
networks reveal how the driving genes are connected.

```{r}
#| label: string-network

string_db <- STRINGdb$new(
  version         = "12.0",
  species         = 10090,   # Mus musculus
  score_threshold = 400,
  network_type    = "full"
)

# Example: Cluster 4 (epigenetic regulation)
genes_cl4 <- cluster_df$gene[cluster_df$div_cluster == 4]
mapped_cl4 <- string_db$map(data.frame(gene = genes_cl4),
                             "gene",
                             removeUnmappedRows = TRUE)

cat(sprintf("Cluster 4: %d / %d genes mapped to STRING\n",
            nrow(mapped_cl4), length(genes_cl4)))

string_db$plot_network(mapped_cl4$STRING_id)
```

---

## Session Info

```{r}
#| label: session-info

sessionInfo()
```
