---
title: "AML Monte Carlo Parameterization — c₃ Prediction Calibration"
format: html
author: Danya Hassan
knitr:
  opts_chunk:
    echo: TRUE
    warning: FALSE
    error: TRUE
---

## Overview

This document calibrates and evaluates the Monte Carlo (MC) simulation used to
predict time-to-leukemia (c₃ first-passage time) for CM mice. The quasi-potential
constructed in `aml2018_state_space_construction.qmd` is first rescaled so that
the energy barrier matches the reference paper value (0.99 A.U.). A
Metropolis–Hastings random walk is then used to simulate transcriptomic
trajectories across the landscape.

Two parameter sweeps are performed:

1. **Step-size sweep** — `step_sd` × `max_mult` grid (18 combinations) to
   identify the proposal distribution that minimises prediction MAE.
2. **Diffusion sweep** — sensitivity of predictions to D across 0.25–0.75,
   retaining D = 0.45 from the reference paper.

Calibration mice: 3334, 3336, 3341, 3357 | Validation mice: 3370, 3338

---

## 1. Setup

```{r}
#| label: libraries

library(ggplot2)
library(dplyr)
library(tidyr)
library(tidyverse)
library(patchwork)
library(readr)
library(tictoc)
```

```{r}
#| label: load-data

mrna_2018  <- readRDS("AML_mRNA_2018_1_real.rds")

aml_ss     <- read.csv("final_aml_mrna_state_space.csv")
aml_qp_dw  <- read.csv("aml_qp_dw_k2max.csv")
all_2018_ss <- read_csv("AML_2018_StateSpace_Mice.csv")
```

```{r}
#| label: plot-trajectories
#| fig-width: 8
#| fig-height: 5

ggplot(aml_ss, aes(x = Time, y = AML_space.calc,
                   colour = factor(mouse_id), group = mouse_id)) +
  geom_point() + geom_line() +
  theme_bw(base_size = 16) +
  xlab("Time (weeks)") + ylab("AML State-Space (PC2)") +
  ggtitle("AML and Ctrl Mice: State-Space Projection")
```

---

## 2. State Boundaries and Landscape Diagnostics

Define the two boundaries that divide the PC2 axis into three discrete macro-states:

- **State 1** — perturbed hematopoiesis (near-healthy)
- **State 2** — transition region
- **State 3** — deep leukemia

```{r}
#| label: state-boundaries

c1 <- 14.89216
c2 <- -40.5698567718201
c3 <- -299.0184

boundary_1 <- (c1 + c2) / 2   # midpoint between c1 and c2
boundary_2 <- -250              # empirical lower boundary for State 3

cat(sprintf("Boundary 1 (S1|S2): %.4f\n", boundary_1))
cat(sprintf("Boundary 2 (S2|S3): %.4f\n", boundary_2))

# Assign discrete states to landscape and sample data
aml_qp_dw$state <- cut(
  aml_qp_dw$AML_PC2_Location,
  breaks = c(-Inf, boundary_2, boundary_1, Inf),
  labels = c("3", "2", "1")
) |> as.character() |> as.integer()

aml_ss$state <- cut(
  aml_ss$AML_space.calc,
  breaks = c(-Inf, boundary_2, boundary_1, Inf),
  labels = c("3", "2", "1")
) |> as.character() |> as.integer()

cat("Landscape state counts:\n"); print(table(aml_qp_dw$state))
cat("Sample state counts:\n");    print(table(aml_ss$state))

dx <- diff(sort(aml_qp_dw$AML_PC2_Location))
cat("\nGrid spacing summary:\n"); print(summary(dx))
cat(sprintf("\nPre-rescaling U range: %.4f to %.4f\n",
            min(aml_qp_dw$U_vals), max(aml_qp_dw$U_vals)))
```

---

## 3. Simulation Parameters and Potential Energy Rescaling

The quasi-potential is rescaled so that the energy barrier at the S1/S2 boundary
equals 0.99 A.U., matching the value reported in the reference paper. All
downstream simulations use this rescaled landscape.

```{r}
#| label: rescale-potential

# Sort landscape by PC2 coordinate
landscape <- aml_qp_dw[order(aml_qp_dw$AML_PC2_Location), ]
x_seq  <- landscape$AML_PC2_Location
U_seq  <- landscape$U_vals
states <- landscape$state
n_pts  <- length(x_seq)

# Rescale U so that the barrier height = 0.99 A.U.
target_barrier  <- 0.99
current_barrier <- U_seq[which.min(abs(x_seq - boundary_1))] -
                   min(U_seq[states == 1L])
u_rescale_factor <- target_barrier / current_barrier
U_seq            <- U_seq * u_rescale_factor

barrier <- U_seq[which.min(abs(x_seq - boundary_1))] -
           min(U_seq[states == 1L])

cat(sprintf("U rescale factor:         %.4f\n", u_rescale_factor))
cat(sprintf("Post-rescaling barrier:   %.4f A.U.\n", barrier))
cat(sprintf("New U range:              %.4f to %.4f\n",
            min(U_seq), max(U_seq)))

# Simulation hyperparameters
D              <- 0.45    # diffusion coefficient (paper value)
n_steps        <- 50000
step_sd_coord  <- 15
max_step_coord <- 22.5

cat(sprintf("\nD:              %.2f\n",  D))
cat(sprintf("D / barrier:    %.4f\n",   D / barrier))
cat(sprintf("State 1: %d pts | State 2: %d pts | State 3: %d pts\n",
            sum(states == 1L), sum(states == 2L), sum(states == 3L)))
cat(sprintf("State 1 coord range: [%.2f, %.2f]\n",
            min(x_seq[states == 1L]), max(x_seq[states == 1L])))
```

---

## 4. Metropolis–Hastings Proposal and FPT Functions

The simulation uses a Gaussian proposal with reflecting boundaries. A move is
accepted with probability min(1, exp(−ΔU / D)). First-passage time (FPT) is
recorded when a trajectory first enters State 3.

```{r}
#| label: fpt-functions

# Proposal function — Gaussian step with reflecting boundaries
propose_index <- function(cur_idx) {
  cur_coord  <- x_seq[cur_idx]
  delta      <- rnorm(1, 0, step_sd_coord)
  delta      <- max(min(delta, max_step_coord), -max_step_coord)
  if (abs(delta) < 1e-6)
    delta <- sample(c(-1, 1), 1) * step_sd_coord * 0.1
  prop_coord <- cur_coord + delta
  x_min <- x_seq[1]; x_max <- x_seq[n_pts]
  if (prop_coord < x_min) prop_coord <- 2 * x_min - prop_coord
  if (prop_coord > x_max) prop_coord <- 2 * x_max - prop_coord
  which.min(abs(x_seq - prop_coord))
}

# FPT via propose_index (uses global step_sd / max_step)
run_fpt <- function(start_idx, n_steps, D) {
  cur <- start_idx
  for (t in seq_len(n_steps)) {
    prop   <- propose_index(cur)
    deltaU <- U_seq[prop] - U_seq[cur]
    if (runif(1) < min(1, exp(-deltaU / D))) cur <- prop
    if (states[cur] == 3L) return(t)
  }
  return(NA_integer_)
}

# FPT with explicit step parameters (used in sweep)
run_fpt_sweep <- function(start_idx, n_steps, D,
                          step_sd_coord, max_step_coord) {
  cur_idx <- start_idx
  for (i in seq_len(n_steps)) {
    cur_coord  <- x_seq[cur_idx]
    delta      <- rnorm(1, 0, step_sd_coord)
    delta      <- max(min(delta, max_step_coord), -max_step_coord)
    if (abs(delta) < 1e-6)
      delta <- sample(c(-1L, 1L), 1L) * step_sd_coord * 0.1
    prop_coord <- cur_coord + delta
    x_min <- x_seq[1L]; x_max <- x_seq[n_pts]
    if (prop_coord < x_min) prop_coord <- 2 * x_min - prop_coord
    if (prop_coord > x_max) prop_coord <- 2 * x_max - prop_coord
    prop_idx <- which.min(abs(x_seq - prop_coord))
    dU <- U_seq[prop_idx] - U_seq[cur_idx]
    if (dU <= 0 || runif(1) < exp(-dU / D)) cur_idx <- prop_idx
    if (x_seq[cur_idx] <= boundary_2) return(i)
  }
  return(NA_integer_)
}

# Pool of starting indices (all State 1 grid points)
c1_idx <- which(states == 1L)
cat(sprintf("Starting pool (State 1): %d points\n", length(c1_idx)))
```

---

## 5. Calibration — step_to_month Conversion

`step_to_month` converts MC steps to calendar months by anchoring the mean
simulated FPT to the mean observed leukemia time across calibration mice.

```{r}
#| label: observed-times

# Calibration / validation split
calibration_mice <- c(3334, 3357, 3336, 3341)
validation_mice  <- c(3370, 3338)

# Observed first-passage times (months)
obs_all <- tibble(
  mouse_id      = c(3334, 3336, 3338, 3341, 3357, 3370),
  observed_time = c(6.5,  7.0,  3.0,  6.0,  3.0,  9.0)
)

# Mean observed time for calibration mice only
mean_obs_months_calib <- mean(
  obs_all$observed_time[obs_all$mouse_id %in% calibration_mice]
)
cat(sprintf("Mean observed leukemia time (calibration mice): %.3f months\n",
            mean_obs_months_calib))
```

```{r}
#| label: starting-coordinates

# Each mouse starts at its first observed timepoint coordinate
start_coords <- aml_ss %>%
  filter(group_name == "CM") %>%
  arrange(mouse_id, Time) %>%
  group_by(mouse_id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  select(mouse_id, AML_space.calc)

start_coords %>%
  left_join(obs_all, by = "mouse_id") %>%
  mutate(role = if_else(mouse_id %in% calibration_mice,
                        "calibration", "validation")) %>%
  arrange(observed_time)
```

```{r}
#| label: calibrate-step-to-month

set.seed(42)

calib_steps <- vapply(seq_len(200), function(i)
  run_fpt(sample(c1_idx, 1), n_steps, D), integer(1))

na_frac <- mean(is.na(calib_steps))
cat(sprintf("Calibration: %.0f%% reached State 3 within %d steps\n",
            (1 - na_frac) * 100, n_steps))
if (na_frac > 0.2)
  warning("More than 20% NA — consider increasing n_steps.")

mean_calib_steps <- mean(calib_steps, na.rm = TRUE)
step_to_month    <- mean_obs_months_calib / mean_calib_steps

cat(sprintf("Mean calibration steps: %.1f\nstep_to_month:          %.6f\n",
            mean_calib_steps, step_to_month))
```

---

## 6. Step-Size Parameter Sweep

A 6 × 3 grid of (`step_sd`, `max_mult`) combinations is evaluated. For each
combination `step_to_month` is recalibrated independently to ensure a fair MAE
comparison. Combinations with < 80% of trajectories reaching State 3 are
excluded.

```{r}
#| label: step-sweep

step_sd_vals  <- c(2, 5, 10, 15, 20, 30)
max_mult_vals <- c(1.5, 2.0, 3.0)
sweep_grid    <- expand.grid(step_sd = step_sd_vals, max_mult = max_mult_vals)

n_reps        <- 700L
n_steps_sweep <- 20000L

sweep_results <- vector("list", nrow(sweep_grid))

for (g in seq_len(nrow(sweep_grid))) {

  step_sd_g  <- sweep_grid$step_sd[g]
  max_step_g <- sweep_grid$step_sd[g] * sweep_grid$max_mult[g]

  set.seed(42)
  calib_steps_g   <- vapply(seq_len(200), function(i)
    run_fpt_sweep(sample(c1_idx, 1), n_steps_sweep, D,
                  step_sd_g, max_step_g), integer(1))
  step_to_month_g <- mean_obs_months_calib / mean(calib_steps_g, na.rm = TRUE)

  cat(sprintf("Combo %02d/18 | step_sd=%-3g max_mult=%.1f | step_to_month=%.6f\n",
              g, step_sd_g, sweep_grid$max_mult[g], step_to_month_g))

  combo_rows <- lapply(seq_len(nrow(start_coords)), function(m) {
    mid       <- start_coords$mouse_id[m]
    start_idx <- which.min(abs(x_seq - start_coords$AML_space.calc[m]))

    fpt_steps  <- vapply(seq_len(n_reps), function(i)
      run_fpt_sweep(start_idx, n_steps_sweep, D,
                    step_sd_g, max_step_g), integer(1))
    fpt_months <- fpt_steps * step_to_month_g
    obs_t      <- obs_all$observed_time[obs_all$mouse_id == mid]

    tibble(
      step_sd   = step_sd_g,
      max_mult  = sweep_grid$max_mult[g],
      mouse_id  = mid,
      pred_mean = mean(fpt_months,   na.rm = TRUE),
      pred_med  = median(fpt_months, na.rm = TRUE),
      obs_time  = obs_t,
      pct_s3    = mean(!is.na(fpt_steps)) * 100,
      ae_mean   = abs(mean(fpt_months,   na.rm = TRUE) - obs_t),
      ae_med    = abs(median(fpt_months, na.rm = TRUE) - obs_t)
    )
  })
  sweep_results[[g]] <- bind_rows(combo_rows)
}

sweep_df <- bind_rows(sweep_results)

sweep_summary <- sweep_df %>%
  group_by(step_sd, max_mult) %>%
  summarise(
    MAE_mean   = mean(ae_mean, na.rm = TRUE),
    MAE_median = mean(ae_med,  na.rm = TRUE),
    pct_s3_avg = mean(pct_s3,  na.rm = TRUE),
    .groups = "drop"
  )

best_combo_mean   <- sweep_summary %>%
  filter(pct_s3_avg >= 80) %>%
  slice_min(MAE_mean, n = 1)
best_combo_median <- sweep_summary %>%
  filter(pct_s3_avg >= 80) %>%
  slice_min(MAE_median, n = 1)

cat(sprintf("\nBest by mean MAE:   step_sd=%.0f, max_mult=%.1f | MAE=%.3f | pct_s3=%.1f%%\n",
            best_combo_mean$step_sd, best_combo_mean$max_mult,
            best_combo_mean$MAE_mean, best_combo_mean$pct_s3_avg))
cat(sprintf("Best by median MAE: step_sd=%.0f, max_mult=%.1f | MAE=%.3f | pct_s3=%.1f%%\n",
            best_combo_median$step_sd, best_combo_median$max_mult,
            best_combo_median$MAE_median, best_combo_median$pct_s3_avg))
```

### 6a. Step-sweep diagnostic plots

```{r}
#| label: plot-step-sweep
#| fig-width: 9
#| fig-height: 6

# Median MAE heatmap
p_heatmap_med <- ggplot(sweep_summary,
    aes(x = factor(step_sd), y = factor(max_mult), fill = MAE_median)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.2f", MAE_median)),
            size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#1D9E75", high = "#D85A30",
                      name = "Median MAE\n(months)") +
  labs(title    = "Median MAE by step_sd and max_step multiplier",
       subtitle = "Green = lower error (better)",
       x = "step_sd (PC2 units)",
       y = "max_step multiplier (× step_sd)") +
  theme_bw(base_size = 13)

# Mean MAE heatmap
p_heatmap_mean <- ggplot(sweep_summary,
    aes(x = factor(step_sd), y = factor(max_mult), fill = MAE_mean)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.2f", MAE_mean)),
            size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#1D9E75", high = "#D85A30",
                      name = "Mean MAE\n(months)") +
  labs(title    = "Mean MAE by step_sd and max_step multiplier",
       subtitle = "Green = lower error (better)",
       x = "step_sd (PC2 units)",
       y = "max_step multiplier (× step_sd)") +
  theme_bw(base_size = 13)

# % reaching State 3
p_reach <- ggplot(sweep_summary,
    aes(x = factor(step_sd), y = factor(max_mult), fill = pct_s3_avg)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.0f%%", pct_s3_avg)), size = 4) +
  scale_fill_gradient(low = "#D85A30", high = "#1D9E75",
                      limits = c(0, 100),
                      name = "% reached\nstate 3") +
  labs(title    = "% trajectories reaching State 3",
       subtitle = "< 80% → n_steps too short or step size too small",
       x = "step_sd (PC2 units)",
       y = "max_step multiplier (× step_sd)") +
  theme_bw(base_size = 13)

# Median MAE vs step_sd line plot
p_line <- ggplot(sweep_summary,
    aes(x = step_sd, y = MAE_median,
        color = factor(max_mult), group = factor(max_mult))) +
  geom_line(linewidth = 1) + geom_point(size = 3) +
  scale_color_brewer(palette = "Set1", name = "max_step\nmultiplier") +
  labs(title = "Median MAE vs step_sd by max_step multiplier",
       x = "step_sd (PC2 units)", y = "Median MAE (months)") +
  theme_bw(base_size = 13)

print(p_heatmap_med)
print(p_heatmap_mean)
print(p_reach)
print(p_line)
```

```{r}
#| label: plot-best-combo
#| fig-width: 6
#| fig-height: 5

# Per-mouse scatter for the best combo (step_sd = 15, max_mult = 1.5)
best_mouse_df <- sweep_df %>%
  filter(step_sd == 15, max_mult == 1.5) %>%
  mutate(role = if_else(mouse_id %in% calibration_mice,
                        "calibration", "validation"))

p_best <- ggplot(best_mouse_df,
    aes(x = obs_time, y = pred_med, color = role, label = mouse_id)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey50") +
  geom_point(size = 4) +
  ggrepel::geom_label_repel(size = 3.5) +
  scale_color_manual(values = c("calibration" = "#3B8BD4",
                                "validation"  = "#D85A30")) +
  labs(title    = "Best combo: step_sd = 15, max_mult = 1.5",
       subtitle = "Predicted median vs observed. Dashed = perfect prediction.",
       x = "Observed time (months)", y = "Predicted median (months)") +
  theme_bw(base_size = 13)

print(p_best)
```

> **Selected parameters: step_sd = 15, max_mult = 1.5.**
> This combination achieves the lowest mean MAE among all combinations with
> ≥ 80% State-3 reachability, and represents a physically reasonable proposal
> step size (~18× the grid spacing). The step_sd = 2 advantage on median MAE
> is likely an artifact of recalibration compressing many small steps into a
> narrow predicted-time range.

---

## 7. D Sensitivity Sweep

`step_to_month` is recalibrated at each D value. Validation mice (3370, 3338)
are held out from D selection throughout.

```{r}
#| label: d-sweep

D_vals       <- seq(0.25, 0.75, by = 0.10)
n_sweep_reps <- 200L

cat("Running D sensitivity sweep...\n")

d_sweep_results <- lapply(D_vals, function(d) {
  cat(sprintf("  D = %.2f\n", d))

  # Recalibrate at this D
  calib_steps_d <- vapply(seq_len(200), function(i)
    run_fpt_sweep(sample(c1_idx, 1), n_steps, d,
                  step_sd_coord, max_step_coord), integer(1))
  step_to_month_d <- mean_obs_months_calib / mean(calib_steps_d, na.rm = TRUE)

  mouse_rows <- lapply(seq_len(nrow(start_coords)), function(m) {
    mid       <- start_coords$mouse_id[m]
    start_idx <- which.min(abs(x_seq - start_coords$AML_space.calc[m]))

    steps <- vapply(seq_len(n_sweep_reps), function(i)
      run_fpt_sweep(start_idx, n_steps, d,
                    step_sd_coord, max_step_coord), integer(1))

    sim_months  <- steps * step_to_month_d
    obs_t       <- obs_all$observed_time[obs_all$mouse_id == mid]

    tibble(
      D             = d,
      mouse_id      = mid,
      start_coord   = start_coords$AML_space.calc[m],
      step_to_month = step_to_month_d,
      pct_s3        = mean(!is.na(steps)) * 100,
      pred_mean     = mean(sim_months,   na.rm = TRUE),
      pred_median   = median(sim_months, na.rm = TRUE),
      obs_time      = obs_t,
      ae_mean       = abs(mean(sim_months,   na.rm = TRUE) - obs_t),
      ae_median     = abs(median(sim_months, na.rm = TRUE) - obs_t),
      role          = if_else(mid %in% calibration_mice,
                              "calibration", "validation")
    )
  })
  bind_rows(mouse_rows)
})

d_sweep_df <- bind_rows(d_sweep_results)

d_sweep_summary <- d_sweep_df %>%
  group_by(D) %>%
  summarise(
    mae_mean_all     = mean(ae_mean,                          na.rm = TRUE),
    mae_median_all   = mean(ae_median,                        na.rm = TRUE),
    mae_mean_calib   = mean(ae_mean[role == "calibration"],   na.rm = TRUE),
    mae_mean_val     = mean(ae_mean[role == "validation"],    na.rm = TRUE),
    mae_median_calib = mean(ae_median[role == "calibration"], na.rm = TRUE),
    mae_median_val   = mean(ae_median[role == "validation"],  na.rm = TRUE),
    pct_s3_avg       = mean(pct_s3,                          na.rm = TRUE),
    step_to_month    = first(step_to_month),
    .groups = "drop"
  )

best_D <- d_sweep_summary %>%
  filter(pct_s3_avg >= 80) %>%
  slice_min(mae_median_calib, n = 1)

cat("\n=== D Sensitivity Sweep Summary ===\n")
print(d_sweep_summary)
cat(sprintf("\nBest D (by median MAE on calibration mice): %.2f\n", best_D$D))
cat(sprintf("  MAE_median(calib) = %.3f | MAE_mean(calib) = %.3f | pct_s3 = %.1f%%\n",
            best_D$mae_median_calib, best_D$mae_mean_calib, best_D$pct_s3_avg))
```

### 7a. D sweep diagnostic plots

```{r}
#| label: plot-d-sweep-aggregate
#| fig-width: 8
#| fig-height: 10

# Plot 1: Overall MAE (mean + median) vs D
p_d1 <- d_sweep_summary %>%
  select(D, mae_mean_all, mae_median_all) %>%
  pivot_longer(-D, names_to = "stat", values_to = "mae") %>%
  ggplot(aes(x = D, y = mae, color = stat)) +
  geom_line(linewidth = 1) + geom_point(size = 3) +
  geom_vline(xintercept = 0.45, linetype = "dashed", color = "grey40") +
  annotate("text", x = 0.46, y = max(d_sweep_summary$mae_mean_all) * 0.95,
           label = "D = 0.45\n(paper)", hjust = 0, size = 3.2, color = "grey40") +
  scale_color_manual(values = c("mae_mean_all"   = "#7F77DD",
                                "mae_median_all" = "#1D9E75"),
                     labels = c("Mean MAE", "Median MAE")) +
  labs(title = "MAE vs D (all mice)",
       x = "D", y = "MAE (months)", color = NULL) +
  theme_bw(base_size = 13)

# Plot 2: Median MAE — calibration vs validation
p_d2 <- d_sweep_summary %>%
  select(D, mae_median_calib, mae_median_val) %>%
  pivot_longer(-D, names_to = "group", values_to = "mae") %>%
  ggplot(aes(x = D, y = mae, color = group)) +
  geom_line(linewidth = 1) + geom_point(size = 3) +
  geom_vline(xintercept = best_D$D, linetype = "dashed", color = "grey40") +
  annotate("text", x = best_D$D + 0.01,
           y = max(d_sweep_summary$mae_median_val) * 0.97,
           label = sprintf("Best D = %.2f", best_D$D),
           hjust = 0, size = 3.2, color = "grey40") +
  scale_color_manual(values = c("mae_median_calib" = "#534AB7",
                                "mae_median_val"   = "#D85A30"),
                     labels = c("Calibration mice", "Validation mice")) +
  labs(title    = "Median MAE vs D — calibration vs validation",
       subtitle = "Validation mice (3370, 3338) never used to select D",
       x = "D", y = "Median MAE (months)", color = NULL) +
  theme_bw(base_size = 13)

# Plot 3: Mean MAE — calibration vs validation
p_d3 <- d_sweep_summary %>%
  select(D, mae_mean_calib, mae_mean_val) %>%
  pivot_longer(-D, names_to = "group", values_to = "mae") %>%
  ggplot(aes(x = D, y = mae, color = group)) +
  geom_line(linewidth = 1) + geom_point(size = 3) +
  geom_vline(xintercept = best_D$D, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = c("mae_mean_calib" = "#534AB7",
                                "mae_mean_val"   = "#D85A30"),
                     labels = c("Calibration mice", "Validation mice")) +
  labs(title    = "Mean MAE vs D — calibration vs validation",
       subtitle = "Validation mice (3370, 3338) never used to select D",
       x = "D", y = "Mean MAE (months)", color = NULL) +
  theme_bw(base_size = 13)

print(p_d1)
print(p_d2 / p_d3)
```

```{r}
#| label: plot-d-sweep-per-mouse
#| fig-width: 8
#| fig-height: 10

# Plot 4: Per-mouse predicted median vs D
p_d4 <- ggplot(d_sweep_df,
    aes(x = D, y = pred_median,
        color = factor(mouse_id), linetype = role)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  geom_hline(data = obs_all,
             aes(yintercept = observed_time, color = factor(mouse_id)),
             linetype = "dotted", linewidth = 0.7) +
  scale_linetype_manual(values = c("calibration" = "solid",
                                   "validation"  = "dashed")) +
  labs(title    = "Per-mouse predicted median vs D",
       subtitle = "Dotted horizontal = observed leukemia time",
       x = "D", y = "Predicted median (months)",
       color = "Mouse", linetype = "Role") +
  theme_bw(base_size = 13)

# Plot 5: Per-mouse predicted mean vs D
p_d5 <- ggplot(d_sweep_df,
    aes(x = D, y = pred_mean,
        color = factor(mouse_id), linetype = role)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  geom_hline(data = obs_all,
             aes(yintercept = observed_time, color = factor(mouse_id)),
             linetype = "dotted", linewidth = 0.7) +
  scale_linetype_manual(values = c("calibration" = "solid",
                                   "validation"  = "dashed")) +
  labs(title    = "Per-mouse predicted mean vs D",
       subtitle = "Dotted horizontal = observed leukemia time",
       x = "D", y = "Predicted mean (months)",
       color = "Mouse", linetype = "Role") +
  theme_bw(base_size = 13)

print(p_d4 / p_d5)
```

```{r}
#| label: plot-d-sweep-ae-faceted
#| fig-width: 9
#| fig-height: 7

# Plot 6: Per-mouse AE (mean + median) vs D — faceted
p_d6 <- d_sweep_df %>%
  select(D, mouse_id, role, ae_mean, ae_median) %>%
  pivot_longer(cols = c(ae_mean, ae_median),
               names_to = "stat", values_to = "mae") %>%
  mutate(stat = recode(stat,
                       "ae_mean"   = "Mean AE",
                       "ae_median" = "Median AE")) %>%
  ggplot(aes(x = D, y = mae, color = stat, linetype = role)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  geom_vline(xintercept = best_D$D,
             linetype = "dashed", color = "grey40") +
  scale_color_manual(values = c("Mean AE"   = "#7F77DD",
                                "Median AE" = "#1D9E75")) +
  scale_linetype_manual(values = c("calibration" = "solid",
                                   "validation"  = "dashed")) +
  facet_wrap(~ mouse_id, scales = "free_y") +
  labs(title    = "Per-mouse MAE vs D (faceted)",
       subtitle = "Solid = calibration | dashed = validation | grey dashed = best D",
       x = "D", y = "Absolute Error (months)",
       color = "Statistic", linetype = "Role") +
  theme_bw(base_size = 12)

print(p_d6)
```

```{r}
#| label: plot-d-reachability
#| fig-width: 6
#| fig-height: 4

# Plot 7: % reaching State 3 vs D
p_d7 <- ggplot(d_sweep_summary, aes(x = D, y = pct_s3_avg)) +
  geom_line(color = "#1D9E75", linewidth = 1) +
  geom_point(color = "#1D9E75", size = 3) +
  geom_hline(yintercept = 80, linetype = "dashed", color = "#D85A30") +
  annotate("text", x = 0.26, y = 81.5,
           label = "80% threshold", hjust = 0, size = 3.2, color = "#D85A30") +
  labs(title    = "% trajectories reaching State 3 vs D",
       subtitle = "Below 80% → n_steps too short for this D",
       x = "D", y = "% reached State 3") +
  theme_bw(base_size = 13)

print(p_d7)
```

---

## 8. D Selection — Rationale

```{r}
#| label: d-mae-summary

d_sweep_summary %>%
  select(D, mae_mean_all, mae_median_all, pct_s3_avg) %>%
  arrange(mae_mean_all)
```

> **Selected value: D = 0.45 (paper value).**
>
> MAE varied by less than 0.5 months across D = 0.25–0.75, with no clear
> minimum on the calibration set. This is consistent with all six mice starting
> within a ~20 PC2-unit window in a near-flat region of the landscape, where
> individual starting position explains only ~29% of variance in observed
> leukemia time. D = 0.45 performs near-optimally on this cohort and was
> validated on a substantially larger training cohort in the reference paper.
> Deviating from it to chase a sub-0.2-month MAE improvement on four
> calibration mice would constitute overfitting.

---

## Session Info

```{r}
#| label: session-info

sessionInfo()
```
